\documentclass[a4paper]{article}

\def\npart {III}
\def\nterm {Easter}
\def\nyear {2017}
\def\nlecturer {N.\ S.\ Manton and D.\ Stuart}
\def\ncourse {Classical and Quantum Solitons}
\def\nnotready {}

\input{header}

\usepackage[compat=1.1.0]{tikz-feynman}
\tikzfeynmanset{/tikzfeynman/momentum/arrow shorten = 0.3}
\tikzfeynmanset{/tikzfeynman/warn luatex = false}


\begin{document}
\maketitle
{\small
\setlength{\parindent}{0em}
\setlength{\parskip}{1em}
Solitons are solutions of classical field equations with particle-like properties. They are localised in space, have finite energy and are stable against decay into radiation. The stability usually has a topological explanation. After quantisation, they give rise to new particle states in the underlying quantum field theory that are not seen in perturbation theory. We will focus mainly on kink solitons in one space dimension, and on Skyrmions in three dimensions. Solitons in gauge theories will also be mentioned.

\subsubsection*{Pre-requisites}
This course assumes you have taken Quantum Field Theory and Symmetries, Fields and Particles. The small amount of topology that is needed will be developed during the course.
}
\tableofcontents

\setcounter{section}{-1}
\section{Introduction}
Given a classical field theory, if we want to ``quantize'' it, then we find the vacuum of the theory, and then do perturbation theory around this vacuum. If there are multiple vacua, then what we did was that we arbitrarily picked a vacuum, and then expanded around that vacuum.

However, these field theories with multiple vacua often contain \emph{soliton} solutions. These are localized, smooth solutions of the classical field equations, and they ``connect multiple vacuums''. To quantize these solitons solutions, we fix such a soliton, and use it as the ``background''. We then do perturbation theory around these solutions, but this is rather tricky to do. Thus, in a lot of the course, we will just look at the classical part of the theory.

Recall that when quantizing our field theories in perturbation theory, we obtain particles in the quantum theory, despite the classical theory being completely about fields. It turns our solitons also behave like particles, and they are a \emph{new} type of particles. These are non-perturbative phenomenon. If we want to do the quantum field theory properly, we have to include these solitons in the quantum field theory. In general this is hard, and so we are not going to develop this a lot.

What does it mean to say that solitons are like particles? In relativistic field theories, we find these solitons have a classical energy. We define the ``mass'' $M$ of the soliton to be the energy in the ``rest frame''. Since this is relativistic, we can do a Lorentz boost, and we obtain a moving soliton. Then we obtain a relation of the form
\[
  E^2 - \mathbf{P} \cdot \mathbf{P} = M^2.
\]
This is a Lorentz-invariant property of the soliton. Together with the fact that the soliton is localized, this is a justification for thinking of them as particles.

These particles differ from the particles perturbative quantum fields, as they have rather different properties. Interesting solitons have a \emph{topological} character different from the classical vacuum. Thus, at least naively, they cannot be thought of perturbatively.

There are also non-relativistic solitons, and they usually don't have interpretations of particles. These appear, for example, as defects in solids. We will not be interested in these much.

What kinds of theories have solitons? To obtain solitons, we definitely need a non-linear field structure and/or non-linear equations. Thus, free field theories with quadratic Lagrangians such as Maxwell theory do not have solitons. We need interaction terms.

Note that in QFT, we did interactions using the interaction picture. We split the Hamiltonian into a ``free field'' part, which we solve exactly, and the ``interaction'' part. However, to quantize solitons, we need to solve the full interacting Lagrangian \emph{exactly}.

Having interactions is not enough for solitons to appear. To obtain solitons, we also need some non-trivial vacuum topology. In other words, we need more than one vacuum. This usually comes from symmetry breaking, and often gauge symmetries are involved.

In this course, we will focus on three types of solitons.
\begin{itemize}
  \item In one (space) dimension, we have kinks. We will spend $4$ lectures on this.

  \item In two dimensions, we have vortices. We will spend $6$ lectures on this.

  \item In three dimensions, there are monopoles and Skyrmions. We will only study Skyrmions, and will spend $6$ lectures on it.
\end{itemize}
These examples are all relativistic. Non-relativistic solitons include \emph{domain walls}, which occur in ferromagnets, but we will not study these.

In general, solitons appear in all sorts of different actual, physical scenarios such as in condensed matter physics, optical fibers, superconductors, ``cosmic strings'' etc. Since we are mathematicians, we probably will not put much focus on these actual applications. However, we can talk a bit more about Skyrmions.

Skyrmions are solitons in an \emph{effective field theory} of interacting pions, which are thought to be the most important baryons because they are the lightest. This happens in spite of the lack of a gauge symmetry. While pions have no baryon number, the associated solitons have a topological charge identified with baryon number. This baryon number is conserved for topological reasons.

Note that in QCD, baryon number is conserved because the quark number is conserved. We tried extremely hard to find proton decay, which would be a process that involves baryon numbers, but we cannot find such examples. We have very high experimental certainty that baryon number is conserved. And if baryon number is topological, then this is a very good reason for the conservation of baryon numbers.

Skyrmions give a model of low-energy interactions of baryons. This leads to an (approximate) theory of nucleons (proton and neutron) and larger nuclei, which are bound states of any number of protons and neutrons. % This is really up-to-date stuff.

There is a whole other set of Skyrmions studied, which are two-dimensional. These are structure in exotic magnets, and they have actually been seen.

For these ideas to work out well, we need to eventually do quantization. For example, Skyrmions by themselves do not have spin. We need to quantize the theory before these come out. Also, Skyrmions cannot distinguish between protons and neutrons. These differences only come up after we quantize.
\section{\tph{$\phi^4$}{phi4}{&straightphi;<sup>4</sup>} kinks}
\subsection{Kink solutions}
In this chapter, we are going to study \term{$\phi^4$ kinks}\index{kink}\index{kink!$\phi^4$}. This happens in $1 + 1$ dimensions, and involves a single scalar field $\phi(x, t)$. In higher dimensions, we often need many fields to obtain solitons, but in the case of 1 dimension, we can get away with a single field.

In general, the \term{Lagrangian density} of such a scalar field theory is of the form
\[
  \mathcal{L} = \frac{1}{2} \partial_\mu \phi \partial^\mu \phi - U(\phi)
\]
for some potential $U(\phi)$ polynomial in $\phi$. Note that in $1 + 1$ dimensions, any such theory is renormalizable. Here we will choose the Minkowski metric to be
\[
  \eta^{\mu\nu} =
  \begin{pmatrix}
    1 & 0\\
    0 & -1
  \end{pmatrix}.
\]
Then the \term{Lagrangian} is given by
\[
  L = \int_{-\infty}^\infty \mathcal{L}\;\d x = \int_{-\infty}^\infty \left(\frac{1}{2} \partial_\mu \phi \partial^\mu \phi - U(\phi)\right)\;\d x,
\]
and then the \term{action} is
\[
  S[\phi] = \int L\;\d t = \int \mathcal{L} \;\d x\;\d t.
\]

There is a non-linearity in the field equations due to a potential $U(\phi)$ with \emph{multiple vacua}. We need this multiple vacua to obtain a soliton. The kink stability comes from the \emph{topology}. It is very simple here, and just comes from counting the discrete, distinct vacua.

As usual, we will write
\[
  \dot{\phi} = \frac{\partial \phi}{\partial t},\quad \phi' = \frac{\partial \phi}{\partial x}.
\]
Often it is convenient to (non-relativistically) split the Lagrangian as
\[
  L = T - V,
\]
where
\[
  T = \int \frac{1}{2} \dot{\phi}^2\;\d x,\quad V = \int \left(\frac{1}{2} \phi'^2 + U(\phi)\right)\;\d x.
\]
In higher dimensions, we separate out $\partial_\mu \phi$ into $\dot{\phi}$ and $\nabla \phi$.

The classical field equation comes from the condition that $S[\phi]$ is stationary under variations of $\phi$. By a standard manipulation, the field equation turns out to be
\[
  \partial_\mu \partial^\mu \phi + \frac{\d U}{\d \phi} = 0.
\]
This is an example of a Klein--Gordon type of field equation, but is non-linear if $U$ is not quadratic. This is known as the \term{non-linear Klein--Gordon equation}\index{Klein--Gordon equation!non-linear}.

We are interested in a soliton that is a static solution. For a static field, the time derivatives can be dropped, and this equation becomes
\[
  \frac{\d^2 \phi}{\partial x^2} = \frac{\d U}{\d \phi}.
\]
Of course, the important part is the choice of $U$! In $\phi^4$ theory, we choose
\[
  U(\phi) = \frac{1}{2} (1 - \phi^2)^2.
\]
This is mathematically the simplest version, because we set all coupling constants to $1$. If we want to quantize this (or just want to be more sophisticated), then we should include some parameters. However, this tends to become a complete mess.

The importance of this $U$ is that it has two minima:
\begin{center}
  \begin{tikzpicture}
    \draw [->](-3, -1) -- (3, -1) node [right] {$\phi$};
    \draw [->] (0, -1.5) -- (0, 3) node [above] {$U(\phi)$};

    \draw [mblue, thick, domain=-2.4:2.4, samples=50] plot [smooth] (\x, {4 * ((\x/2)^4 - (\x/2)^2)});

    \node at (-1.414, -1) [below] {$-1$};
    \node at (1.414, -1) [below] {$1$};
  \end{tikzpicture}
\end{center}
The two classical vacuum are
\[
  \phi(x) \equiv 1,\quad \phi(x) \equiv -1.
\]
This is, of course, not the only possible choice. We can, for example, include some parameters and set
\[
  U(\phi) = \lambda(m^2 - \phi^2)^2.
\]
If we are more adventurous, we can talk about a $\phi^6$ theory with
\[
  U(\phi) = \lambda \phi^2 (m^2 - \phi^2)^2.
\]
In this case, we have $3$ minima, instead of $2$. Even braver people can choose
\[
  U(\phi) = 1 - \cos \phi.
\]
This has \emph{infinitely} many minima. The field equation involves a $\sin \phi$ term, and is hence this theory is called the \term{sine-Gordon theory} (a pun on the name Klein--Gordon, of course).

The sine-Gordon theory is a special case. While it seems like the most complicated potential so far, it is actually \term{integrable}. This implies we can find explicit exact solutions involving multiple, interacting solitons in a rather easy way. However, integrable systems is a topic for another course, namely the IID Integrable Systems course.

For now, we will focus on our simplistic $\phi^4$ theory. As mentioned, there are two vacuum field configurations, both of zero energy. We will in general use the term ``\term{field configuration}'' to refer to fields that are not necessarily solutions to the classical field equation, but in this case, the vacua are indeed solutions.

If we wanted to quantize this $\phi^4$ theory, then we have to pick one of the vacua and do perturbation theory around it. This is known as \term{spontaneous symmetry breaking}. Of course, by symmetry, we obtain the same quantum theory regardless of which vacuum we expand around.

However, as we mentioned, when we want to study solitons, we have to involve \emph{both} vacua. We want to consider solutions that ``connect'' these two vacua. In other words, we are looking for solutions that look like
\begin{center}
  \begin{tikzpicture}
    \draw [->] (-3, 0) -- (3, 0) node [right] {$x$};
    \draw [->] (0, -2) -- (0, 2) node [above] {$\phi$};

    \draw [mblue, thick, domain=-3:3] plot [smooth] (\x, {tanh(1.5*(\x + 1))});

    \draw [dashed] (-3, 1) -- (3, 1);
    \draw [dashed] (-3, -1) -- (3, -1);

    \node [circ] at (-1, 0) {};
    \node [anchor = north west] at (-1, 0) {$a$};
  \end{tikzpicture}
\end{center}
This is known as a \term{kink solution}.

To actually find such a solution, we need the full field equation, given by
\[
  \frac{\d^2 \phi}{\d x^2} = -2 (1 - \phi^2) \phi.
\]
Instead of solving this directly, we will find the kink solutions by considering the energy, since this method generalizes better.

We will work with a general potential $U$ with minimum $0$. From Noether's theorem, we obtain a conserved energy
\[
  E = \int \left(\frac{1}{2} \dot{\phi}^2 + \frac{1}{2} \phi'^2 + U(\phi)\right)\;\d x.
\]
For a static field, we drop the $\dot{\phi}^2$ term. Then this is just the $V$ appearing in the Lagrangian. By definition, the field equation tells us the field is a stationary point of this energy. To find the kink solution, we will in fact find a \emph{minimum} of the energy.

Of course, the global minimum is attained when we have a vacuum field, in which case $E = 0$. However, this is just the global minimum only if we don't impose any boundary conditions. In our case, the kinks satisfy the boundary conditions ``$\phi(\infty) = 1$'', ``$\phi(-\infty) = -1$'' (interpreted in terms of limits, of course). The kinks will minimize energy subject to these boundary conditions.

These boundary conditions are important, because they are ``topological''. Eventually, we will want to understand the dynamics of solitons, so we will want to consider fields that evolve with time. From physical considerations, for any fixed $t$, the field $\phi(x, t)$ must satisfy $\phi(x, t) \to \text{vacuum}$ as $x \to \pm \infty$, or else the field will have infinite energy. However, the vacuum of our potential $U$ is discrete. Thus, if $\phi$ were to evolve continuously with time, the boundary conditions must not evolve with time! At least, this is what we expect classically. Who knows what weird tunnelling can happen in quantum field theory.

So from now on, we fix a some boundary conditions $\phi(\infty)$ and $\phi(-\infty)$, and focus on fields that satisfy these boundary conditions. The trick is to write the potential in the form
\[
  U (\phi) = \frac{1}{2} \left(\frac{\d W (\phi)}{\d \phi}\right)^2.
\]
If $U$ is always non-negative, then we can always find $W$ in principle --- we take the square root and then integrate it. However, in practice, this is useful only if we can find a simple form for $W$. Let's assume we've done that. Then we can write
\begin{align*}
  E &= \frac{1}{2} \int \left(\phi'^2 + \left(\frac{\d W}{\d \phi}\right)^2 \right)\;\d x\\
  &= \frac{1}{2} \int \left(\phi' \mp \frac{\d W}{\d \phi}\right)^2\;\d x \pm \int \frac{\d W}{\d \phi} \frac{\d \phi}{\d x} \;\d x\\
  &= \frac{1}{2} \int \left(\phi' \mp \frac{\d W}{\d \phi}\right)^2\;\d x \pm \int \d W\\
  &= \frac{1}{2} \int \left(\phi' \mp \frac{\d W}{\d \phi}\right)^2\;\d x \pm (W(\phi(\infty)) - W(\phi(-\infty))).
\end{align*}
The second term depends purely on the boundary conditions, which we have fixed. Thus, we can minimize energy if we can make the first term vanish! Note that when completing the square, the choice of the signs is arbitrary. However, if we want to set the first term to be $0$, the second term had better be non-positive, since the energy itself is non-negative! Hence, we will pick the sign such that the second term is $\geq 0$, and then the energy is minimized when
\[
  \phi' = \pm \frac{\d W}{\d \phi}.
\]
In this case, we have
\[
  E = \pm (W(\infty) - W(-\infty)).
\]
These are known as the \term{Bogomolny equation} and the \term{Bogomolny energy bound}. Note that if we picked the other sign, then we cannot solve the differential equation $\phi' = \pm \frac{\d W}{\d \phi}$, because we know the energy must be non-negative.

For the $\phi^4$ kink, we have
\[
  \frac{\d W}{\d \phi} = 1 - \phi^2.
\]
So we pick
\[
  W = \phi - \frac{1}{3} \phi^3.
\]
So when $W = \pm 1$, we have $W = \pm \frac{2}{3}$. We need to choose the $+$ sign, and then we know the energy (and hence mass) of the kink is
\[
  E \equiv M = \frac{4}{3}.
\]
We now solve for $\phi$. The equation we have is
\[
  \phi' = 1 - \phi^2.
\]
Rearranging gives
\[
  \frac{1}{1 - \phi^2} \d \phi = \d x,
\]
which we can integrate to give
\[
  \phi (x) = \tanh (x - a).
\]
This $a$ is an arbitrary constant of integration, labelling the intersection with the $x$-axis. We think of this as the ``\term{location}'' of the kink.

Note that there is not a unique solution, which is not unexpected by translation invariance. Instead, the solutions are labeled by a parameter $a$. This is known as a \term{modulus} of the solution. In general, there can be multiple moduli, and the space of all possible values of the moduli is known as the \term{moduli space}. In this case, the moduli space is just $\R$.

Is this solution stable? We obtained this kink solution by minimizing the energy within this topological class of solutions (i.e.\ among all solutions with the prescribed boundary conditions). Since a field cannot change the boundary conditions during evolution, it follows that the kink must be stable.

Are there other soliton solutions to the field equations? The solutions are determined by the boundary conditions. Thus, we can classify all soliton solutions by counting all possible combinations of the boundary conditions. We have, of course, two vacuum solutions $\phi \equiv 1$ and $\phi \equiv -1$. There is also an \term{anti-kink}\index{kink!anti-} solution obtained by inverting the kink:
\[
  \phi(x) = - \tanh (x - b).
\]
This also has energy $\frac{4}{3}$.

\subsection{Dynamic kink}
We now want to look at kinks that move. Given what we have done so far, this is trivial. Our theory is Lorentz invariant, so we simply apply a Lorentz boost. Then we obtain a field
\[
  \phi(x, t) = \tanh \gamma (x - vt),
\]
where, as usual
\[
  \gamma = (1 - v^2)^{-1/2}.
\]
But this isn't it. Notice that for small $v$, we can approximate the solution simply by
\[
  \phi(x, t) = \tanh (x - vt).
\]
This looks like a kink solution with a moduli that varies with time slowly. This is known as the \term{adiabatic} point of view.

More generally, let's consider a ``moving kink'' field
\[
  \phi(x, t) = \tanh (x - a(t))
\]
for some function $a(t)$. In general, this is not a solution to the field equation, but if $\dot{a}$ is small, then it is ``approximately a solution''.

We can now explicitly compute that
\[
  \dot{\phi} = - \frac{\d a}{\d t} \phi'.
\]
Let's consider fields of this type, and look at the Lagrangian of the field theory. The kinetic term is given by
\[
  T = \int \frac{1}{2} \dot{\phi}^2\;\d x = \frac{1}{2} \left(\frac{\d a}{\d t}\right)^2 \int \phi'^2 \;\d x = \frac{1}{2} M \left(\frac{\d a}{\d t}\right)^2.
\]
To derive this result, we had to perform the integral $\int \phi'^2 \;\d x$, and if we do that horrible integral, we will find a value that happens to be equal to $M = \frac{4}{3}$. Of course, this is not a coincidence. We can derive this result from more general principles to see that the result of integration is manifestly $M$.

The remaining part of the Lagrangian is less interesting. Since it does not involve taking time derivatives, the time variation of $a$ is not seen by it, and we simply have a constant
\[
  V = \frac{4}{3}.
\]
Then the original field Lagrangian becomes a particle Lagrangian
\[
  L = \frac{1}{2}M \dot{a}^2 - \frac{4}{3}.
\]

Note that when we first formulated the field theory, the action principle required us to find a field that extremizes the action \emph{among all fields}. However, what we are doing now is to restrict to the set of kink solutions only, and then when we solve the variational problem arising from this Lagrangian, we are extremizing the action among fields of the form $\tanh (x - a(t))$. We can think of this as motion in a ``valley'' in the field configuration space. In general, these solutions will not also extremize the action among all fields. However, as we said, it will do so ``approximately'' if $\dot{a}$ is small.

We can obtain an effective equation of motion
\[
  M \ddot{a} = 0,
\]
which is an equation of motion for the variable $a(t)$ \emph{in the moduli space}.

Of course, the solution is just given by
\[
  a(t) = vt + \mathrm{const},
\]
where $v$ is an arbitrary constant, which we interpret as the velocity. In this formulation, we do not have any restrictions on $v$, because we took the ``non-relativistic approximation''. This approximation breaks down when $v$ is large.

There is a geometric interpretation to this. We can view the equation of motion $M\ddot{a} = 0$ as the \emph{geodesic equation} in the moduli space $\R$, and we can think of the coefficient $M$ as specifying a Riemannian metric on the moduli space. In this case, the metric is (a scalar multiple of) the usual Euclidean metric.

This seems like a complicated way of describing such a simple system, but this picture generalizes to higher-dimensional systems and allows us to analyze multi-soliton dynamics.

%We can find the dynamics of $a(t)$ from the Lagrangian of the field theory. Thus, we are reducing the ``field dynamics'' to the ``particle dynamics''. We have
%\[
% T = \int \frac{1}{2} \dot{\phi}^2\;\d x = \frac{1}{2} \left(\frac{\d a}{\d t}\right)^2 \int \phi'^2 \;\d x = \frac{1}{2} M \left(\frac{\d a}{\d t}\right)^2.
%\]
%The factor of $M$ just comes from doing the (tricky) integral explicitly, but we can also work it out from more general principles to make it manifestly $M$, and this is known as Derrick's theorem.
%
%Thus, the kink behaves like a particle with mass $M$! How about the potential energy? The potential energy is \emph{not} time-dependent. We simply integrate some polynomial of $\phi$ over all $x$, and the shift by $a$ does not make a difference. In this case, we have
%\[
% V = \frac{4}{3}.
%\]
%So we've reduced the field Lagrangian to a particle Lagrangian
%\[
% L = \frac{1}{2}M \dot{a}^2 - \frac{4}{3}.
%\]
%We can think of this as motion in a \term{valley} in the field configuration space. We are drifting in the energy minima in the field configuration space.
%
%This method is powerful, and applies to multi-soliton dynamics in high dimensions. From this, we obtain an effective equation of motion in moduli space
%\[
% M \ddot{a} = 0.
%\]
%This has \emph{geometric} interpretation as geometric motion in the moduli space. The moduli space is just the real line $\R$ with its standard metric. We can think of the coefficient $M$ as a Riemannian metric, which happens to be constant (as a function of $a$) in this case.
%
%Of course, the solution is
%\[
% a(t) = vt + \mathrm{const},
%\]
%where $v$ is an arbitrary constant, namely velocity.
%
%In this approximation, $v$ can be anything, and the approximation does not see the speed of light. However, as we plug this back into the actual field equation, we see that the approximation breaks down when $v$ is large.
%
%This motion in moduli space is not exact, but is accurate in the non-relativistic approximation.
%
%This was all rather trivial in our case of kinks. However, it is also important and allows us to analyze the motion of several solitons in higher dimension.
%
%Note that here we started with a Lagrangian that is quadratic in time derivatives of the field. When we pass on to solitons, we have a term that is quadratic in the time derivative in the moduli, and the coefficients provide a Riemannian geometry on the moduli space.

We might ask ourselves if there are multi-kinks in our theory. There aren't in the $\phi^4$ theory, because we saw that the solutions are classified by the boundary conditions, and there are only that many boundary conditions we've completely enumerated. In more complicated theories like sine-Gordon theory, multiple kinks are possible.

However, while we cannot have two kinks, we can have a kink followed by an anti-kink. This actually lies in the ``vacuum sector'' of the theory, but it still looks like it's made up of kinks and anti-kinks, and it is interesting to study these.

\subsection{Soliton interactions}
We now want to study interactions between kinks and anti-kinks, and see how they cause each other to move. So far, we were able to label the position of the particle by its ``center'' $a$, and thus we can sensibly talk about how this center moves. However, this center is well-defined only in the very special case of a pure kink or anti-kink, where we can use symmetry to identify the center. If there is some perturbation, or if we have a kink and an anti-kink, it is less clear what should be considered the center.

Fortunately, we can still talk about the momentum of the field, even if we don't have a well-defined center. Indeed, since our theory has translation invariance, Noether's theorem gives us a conserved current which is interpreted as the momentum.

In general, for a single scalar field in $1 + 1$ dimensions, the Lagrangian density can be written in the form.
\[
  \mathcal{L} = \frac{1}{2} \partial_\mu \phi \partial^\mu \phi - U(\phi).
\]
Applying Noether's theorem, to the translation symmetry, we obtain the \term{energy-momentum tensor}
\[
  T^\mu_\nu = \frac{\partial \mathcal{L}}{\partial (\partial_\mu \phi)} \partial_\nu \phi - \delta^\mu_\nu \mathcal{L}.
\]
Fixing a time and integrating over all space, we obtain the conserved energy and conserved momentum
The conserved energy is
\begin{align*}
  E &= \int_{-\infty}^\infty T^0\!_0 \;\d x = \int_{-\infty}^\infty \left(\frac{1}{2}\dot{\phi}^2 + \frac{1}{2} \phi'^2 + U(\phi)\right)\;\d x\\
  P &= - \int_{-\infty}^\infty T^0_1 \;\d x = - \int_{-\infty}^\infty \dot{\phi} \phi' \;\d x.
\end{align*}
We now focus on our moving kink in the adiabatic approximation of the $\phi^4$ theory. Then the field is given by
\[
  \phi = \tanh (x - a(t)).
\]
Doing another horrible integral, we find that the momentum is just
\[
  P = M \dot{a}.
\]
This is just as we would expect from a particle with mass $M$!

Now suppose what we have is instead an kink-antikink configuration
\begin{center}
  \begin{tikzpicture}
    \draw [->] (-3, 0) -- (3, 0) node [right] {$x$};
    \draw [->] (0, -2) -- (0, 2) node [above] {$\phi$};

    \draw [mblue, thick] plot [smooth] coordinates {(-3.0,-0.99933) (-2.9,-0.99851) (-2.8,-0.99668) (-2.7,-0.99263) (-2.6,-0.98367) (-2.5,-0.96403) (-2.4,-0.92167) (-2.3,-0.83365) (-2.2,-0.66404) (-2.1,-0.37995) (-2.0,-0.00000) (-1.9,0.37995) (-1.8,0.66404) (-1.7,0.83365) (-1.6,0.92167) (-1.5,0.96403) (-1.4,0.98367) (-1.3,0.99263) (-1.2,0.99668) (-1.1,0.99851) (-1.0,0.99933) (-0.9,0.99970) (-0.8,0.99986) (-0.7,0.99994) (-0.6,0.99997) (-0.5,0.99999) (-0.4,0.99999) (-0.3,1.00000) (-0.2,1.00000) (-0.1,1.00000) (0.0,1.00000) (0.1,1.00000) (0.2,1.00000) (0.3,1.00000) (0.4,0.99999) (0.5,0.99999) (0.6,0.99997) (0.7,0.99994) (0.8,0.99986) (0.9,0.99970) (1.0,0.99933) (1.1,0.99851) (1.2,0.99668) (1.3,0.99263) (1.4,0.98367) (1.5,0.96403) (1.6,0.92167) (1.7,0.83365) (1.8,0.66404) (1.9,0.37995) (2.0,-0.00000) (2.1,-0.37995) (2.2,-0.66404) (2.3,-0.83365) (2.4,-0.92167) (2.5,-0.96403) (2.6,-0.98367) (2.7,-0.99263) (2.8,-0.99668) (2.9,-0.99851) (3.0,-0.99933)};
    % map (\x -> (showFFloat (Just 1) x "", showFFloat (Just 5) (tanh (4 * (x + 2)) - tanh (4 * (x - 2)) - 1) "")) [-3,-2.9..3]

    \draw [dashed] (-3, 1) -- (3, 1);
    \draw [dashed] (-3, -1) -- (3, -1);

    \node [circ] at (-2, 0) {};
    \node [anchor = north west] at (-2, 0) {$-a$};

    \node [circ] at (2, 0) {};
    \node [anchor = north east] at (2, 0) {$a$};
  \end{tikzpicture}
\end{center}

Here we have to make the crucial assumption that our kinks are well-separated. Matters get a lot worse when they get close to each other, and it is difficult to learn anything about them analytically. However, by making appropriate approximations, we can understand well-separated kink-antikink configurations.

When the kink and anti-kink are far away, we can separate out the kink and anti-kink. To do so, we first pick a point $b$ lying in between the kink and the anti-kink:

\begin{center}
  \begin{tikzpicture}
    \draw [->] (-3, 0) -- (3, 0) node [right] {$x$};
    \draw [->] (0, -2) -- (0, 2) node [above] {$\phi$};

    \draw [mblue, thick] plot [smooth] coordinates {(-3.0,-0.99933) (-2.9,-0.99851) (-2.8,-0.99668) (-2.7,-0.99263) (-2.6,-0.98367) (-2.5,-0.96403) (-2.4,-0.92167) (-2.3,-0.83365) (-2.2,-0.66404) (-2.1,-0.37995) (-2.0,-0.00000) (-1.9,0.37995) (-1.8,0.66404) (-1.7,0.83365) (-1.6,0.92167) (-1.5,0.96403) (-1.4,0.98367) (-1.3,0.99263) (-1.2,0.99668) (-1.1,0.99851) (-1.0,0.99933) (-0.9,0.99970) (-0.8,0.99986) (-0.7,0.99994) (-0.6,0.99997) (-0.5,0.99999) (-0.4,0.99999) (-0.3,1.00000) (-0.2,1.00000) (-0.1,1.00000) (0.0,1.00000) (0.1,1.00000) (0.2,1.00000) (0.3,1.00000) (0.4,0.99999) (0.5,0.99999) (0.6,0.99997) (0.7,0.99994) (0.8,0.99986) (0.9,0.99970) (1.0,0.99933) (1.1,0.99851) (1.2,0.99668) (1.3,0.99263) (1.4,0.98367) (1.5,0.96403) (1.6,0.92167) (1.7,0.83365) (1.8,0.66404) (1.9,0.37995) (2.0,-0.00000) (2.1,-0.37995) (2.2,-0.66404) (2.3,-0.83365) (2.4,-0.92167) (2.5,-0.96403) (2.6,-0.98367) (2.7,-0.99263) (2.8,-0.99668) (2.9,-0.99851) (3.0,-0.99933)};

    \draw [dashed] (-3, 1) -- (3, 1);
    \draw [dashed] (-3, -1) -- (3, -1);

    \node [circ] at (-2, 0) {};
    \node [anchor = north west] at (-2, 0) {$-a$};

    \node [circ] at (2, 0) {};
    \node [anchor = north east] at (2, 0) {$a$};

    \draw [dashed] (-0.3, -2) -- (-0.3, 2);
    \node [circ] at (-0.3, 0) {};
    \node [anchor = north east] at (-0.3, 0) {$b$};
  \end{tikzpicture}
\end{center}

The choice of $b$ is arbitrary, but we should choose it so that it is far away from both kinks. We will later see that, at least to first order, the result of our computations does not depend on which $b$ we choose. We will declare that the parts to the left of $b$ belongs to the kink, and the parts to the right of $b$ belong to the anti-kink. Then by integrating the energy-momentum tensor in these two regions, we can obtain the momentum of the kink and the anti-kink separately.

We will focus on the kink only. Its momentum is given by
\[
  P = - \int_{-\infty}^b T^0_1 \;\d x = - \int_{-\infty}^b \dot{\phi} \phi' \;\d x.
\]
Since $T^\mu_\nu$ is conserved, we know $\partial_\mu T^\mu\!_\nu = 0$. So we find
\[
  \frac{\partial}{\partial t} T^0\!_1 + \frac{\partial}{\partial x}T^1\!_1 = 0.
\]
By Newton's second law, the force on the kink is given by the rate of change of the momentum:
\begin{align*}
  \frac{\d}{\d t} P &= -\int_{-\infty}^b \frac{\partial}{\partial t} T^0\!_1\;\d x \\
  &= \int_{-\infty}^b \frac{\partial}{\partial x} T^1\!_1\;\d x\\
  &= \left.T^1\!_1\right|_b\\
  &= \left(-\frac{1}{2} \dot{\phi}^2 - \frac{1}{2} \phi'^2 + U\right)_b
\end{align*}
Note that there is no contribution at the $-\infty$ end because it is vacuum and $T^1\!_1$ vanishes.

But we want to actually work out what this is. To do so, we need to be more precise about what our initial configuration is. In this theory, we can obtain it just by adding a kink to an anti-kink. The obvious guess is that it should be
\[
  \phi(x) \overset{?}{=} \tanh(x + a) - \tanh(x - a),
\]
but this has the wrong boundary condition. It vanishes on both the left and the right. So we actually want to subtract $1$, and obtain
\[
  \phi(x) = \tanh(x + a) - \tanh(x - a) - 1 \equiv \phi_1 + \phi_2 - 1.
\]
Note that since our equation of motion is not linear, this is in general not a genuine solution! Here, this is approximately a solution, because the kink and anti-kink are well-separated. However, there is no hope that this will be anywhere near a solution when the kink and anti-kink are small!

Before we move on to compute $\dot{\phi}$ and $\phi'$ explicitly and plugging numbers in, we first make some simplifications and approximations. First, we restrict our attention to fields that are initially at rest. So we have $\dot{\phi} = 0$ at $t = 0$. Of course, the force will cause the kinks to move, but we shall, for now, ignore what happens when they start moving.

That gets rid of one term. Next, we notice that we only care about the expression when evaluated at $b$. Here we have $\phi_2 - 1 \approx 0$. So we can try to expand the expression to first order in $\phi_2 - 1$ (and hence $\phi_2'$), and this gives
\[
  F = -\frac{1}{2} \left(\phi_1'^2 + U(\phi_1) - \phi_1' \phi_2' + (\phi_2 - 1) \frac{\d U}{\d \phi}(\phi_1)\right)_b.
\]
We have a zeroth order term $\phi_1'^2 + U(\phi_1)$. We claim that this must vanish. One way to see this is that this term corresponds to the force when there is no anti-kink $\phi_2$. Since the kink does not exert a force on itself, this must vanish!

Analytically, we can deduce this from the Bogomolny equation, which says for any kink solution $\phi$, we have
\[
  \phi' = \frac{\d W}{\d \phi}.
\]
It then follows that
\[
  \frac{1}{2} \phi'^2 = \frac{1}{2} \left(\frac{\d W}{\d \phi}\right)^2 = U(\phi).
\]
Alternatively, we can just computing it directly! In any case, convince yourself that it indeed vanishes in your favorite way, and then move on.

Finally, we note that the field equations tell us
\[
  \frac{\d U}{\d \phi}(\phi_1) = \phi_1''.
\]
So we can write the force as
\[
  F = \big(-\phi_1' \phi_2' + (\phi_2 - 1) \phi_1''\big)_b.
\]
That's about all the simplifications we can make without getting our hands dirty. We might think we should plug in the $\tanh$ terms and compute, but that is \emph{too} dirty. Instead, we use asymptotic expressions of kinks and anti-kinks far from their centers. Using the definition of $\tanh$, we have
\[
  \phi_1 = \tanh(x + a) = \frac{1 - e^{-2(x + a)}}{1 + e^{-2(x + a)}} \approx 1 - 2e^{-2(x + a)}.
\]
This is valid for $x \gg -a$, i.e.\ to the right of the kink. The constant factor of $2$ in from of the exponential is called the \term{amplitude} of the tail. We will later see that the $2$ appearing in the exponent has the interpretation of the mass of the field.

For $\phi_2$, take the approximation that $x \ll a$. Then
\[
  \phi_2 - 1= -\tanh(x - a) - 1 \approx -2 e^{2(x - a)}.
\]
We assume that our $b$ satisfies both of these conditions. These are obviously easy to differentiate once or twice. Doing this, we obtain
\[
  -\phi_1' \phi_2' = (-4e^{-2(x + a)})(-4 e^{2(x - a)}) = 16 e^{-4a}.
\]
Note that this is independent of $x$. In the formula, the $x$ will turn into a $b$, and we saw that the force is independent of $b$. Similarly, the other term is
\[
  (\phi_2 - 1) \phi_1'' = (-2e^{2(x - a)}) (-8 e^{-2(x + a)}) = 16 e^{-4a}.
\]
Therefore we find
\[
  F = 32 e^{-4a},
\]
and as promised, this is independent of the precise position of the cutoff $b$ we chose.

We can write this in a slightly more physical form. Our initial configuration was symmetric around the $y$-axis, but in reality, only the separation matters. We write the separation of the pair as $s = 2a$. Then we have
\[
  F = 32 e^{-2s}.
\]
What is the interpretation of the factor of $2$? Recall that our potential was given by
\[
  U(\phi) = \frac{1}{2}(1 - \phi^2)^2.
\]
We can do perturbation theory around one of the vacuums, say $\phi = 1$. Thus, we set $\phi = 1 + \eta$, and then expanding gives us
\[
  U(\eta) = \frac{1}{2} (-2\eta)^2 = \frac{1}{2}m^2 \eta^2,
\]
where $m = 2$. This is the same ``2'' that goes into the exponential in the force.

What about the constant factor of $32$? Recall that when we expanded the kink solution, we saw that the amplitude of the tail was $A = 2$. It turns out if we re-did our theory and put back the different possible parameters, we will find that the force is given by
\[
  F = 2 m^2 A^2 e^{-ms}.
\]
This is an interesting and important phenomenon. The mass $m$ was the \emph{perturbative} mass of the field. It is something we obtain by perturbation theory. However, the same mass appears in the force between the solitons, which are non-perturbative phenomenon!

This is perhaps not too surprising. After all, when we tried to understand the soliton interactions, we took the approximation that $\phi_1$ and $\phi_2$ are close to $1$ at $b$. Thus, we are in some sense perturbing around the vacuum $\phi \equiv 1$.

%The perturbative field theory has a meson of mass $2 \hbar$. You might have not noticed this $\hbar$ when doing quantum field theory, because we set $\hbar = 1$. But this makes sense, because in free field theory, we decompose the quantum field into a lot of harmonic oscillators, and the energy of the harmonic oscillator was $\hbar\omega\left(n + \frac{1}{2}\right)$. SO the $\hbar$ should be here.
%
%However, in our soliton, we do \emph{not} have an $\hbar$. That was not a mistake. We have only been working classically.
%
%The soliton has mass $M = \frac{4}{3}$, and this is much larger than the meson mass. Note that one should choose $\hbar$ to be small for perturbation theory to be justified. However, soliton is non-perturbative and has larger mass than the meson).

We can interpret the force between the kink and anti-kink diagrammatically. From the quantum field theory point of view, we can think of this force as being due to meson exchange, and we can try to invent a Feynman diagram calculus that involves solitons. This is a bit controversial, but at least heuristically, we can introduce new propagators representing solitons, using double lines, and draw the interaction as
\begin{center}
  \begin{tikzpicture}
    \begin{feynman}
      \vertex (i);
      \vertex [right=of i] (m);
      \vertex [above right=of m] (f1);
      \vertex [below right=of m] (f2) {$\bar{K}$};
      \vertex [above left=of i] (s1);
      \vertex [below left=of i] (s2) {$K$};

      \diagram*{
        (i) -- [scalar] (m), % label ``meson''
        (f2) -- [double distance=1.5pt] (m) -- [double distance=1.5pt] (f1), % make these double lines
        (s1) -- [double distance=1.5pt] (i) -- [double distance=1.5pt] (s2),
      };
    \end{feynman}
  \end{tikzpicture}
\end{center}
%This is an effective diagram leading to a Yukawa force in $1 + 1$ dimensions, which decays exponentially with separation $s$.
%
%The amplitude of the tail of the of the soliton kink is $A = 2$. The factor of $32$ in the force ultimately comes from the mass being $m = 2$. More generally, one can show that if we put in parameters into our theory, we have
%\[
% F = 2 m^2 A^2 e^{-ms}.
%\]

So what happens to this soliton? The force we derived was positive. So the kink is made to move to the right. By symmetry, we will expect the anti-kink to move towards the left. They will collide!

What happens when they collide? All our analysis so far assumed the kinks were well-separated, so everything breaks down. We can only understand this phenomena numerically. After doing some numerical simulations, we see that there are two regimes:
\begin{itemize}
  \item If the kinks are moving slowly, then they will annihilate into \emph{meson radiation}.
  \item If the kinks are moving very quickly, then bounce off each other.
\end{itemize}
%We can now study the time-dependence. The kinks will accelerate due to standard Newtonian dynamics, and they will move towards each other. However, the full dynamics of the kink-antikink pair is complicated. When they hit each other, they annihilate, and this happens in the regime where the separation is large. Where does the energy go when they annihilate? The answer is that they annihilate into meson radiation, which we can discover by doing numerical simulations.
%
%Annihilation is what happens if they are initially at rest, or slowly moving. However, at high speed, they happen to bounce of each other (of course, there is still some energy loss to radiation). These are very complicated, and we understand this mostly through numerical simulations.

\subsection{Quantization of kink motion}
We now briefly talk about how to quantize kinks. The most naive way of doing so is pretty straightforward. We use the moduli space approximation, and then we have a very simple kink Lagrangian.
\[
  L = \frac{1}{2} M \dot{a}^2.
\]
This is just a free particle moving in $\R$ with mass $M$. This $a$ is known as the \term{collective coordinate} of the kink. Quantizing a free particle is very straightforward. It is just IB Quantum Mechanics. For completeness, we will briefly outline this procedure.

We first put the system in Hamiltonian form. The conjugate momentum to $a$ is given by
\[
  P = M \dot{a}.
\]
Then the Hamiltonian is given by
\[
  H = P \dot{a} - L = \frac{1}{2M} P^2.
\]
Then to quantize, we replace $P$ by the operator $-i\hbar \frac{\partial}{\partial a}$. In this case, the quantum Hamiltonian is given by
\[
  H = - \frac{\hbar^2}{2M} \frac{\partial^2}{\partial a^2}.
\]
A wavefunction is a function of $a$ and $t$, and this is just ordinary QM for a single particle.

As usual, the stationary states are given by
\[
  \psi(a) = e^{i\kappa a},
\]
and the momentum and energy (eigenvalues) are
\[
  P = \hbar \kappa,\quad H = E = \frac{\hbar^2 \kappa^2}{2M} = \frac{P^2}{2M}.
\]

Is this actually ``correct''? Morally speaking, we really should quantize the complete $1 + 1$ dimensional field theory. What would this look like?

In normal quantum field theory, we consider perturbations around a vacuum solution, say $\phi \equiv 1$, and we obtain mesons. Here if we want to quantize the kink solution, we should consider field oscillations around the kink. Then the solution contains both a kink and a meson. These mesons give rise to quantum corrections to the kink mass $M$.

Should we be worried about these quantum corrections? Unsurprisingly, it turns out these quantum corrections are of the order of the meson mass. So we should not be worried when the meson mass is small.

Meson-Kink scattering can also be studied in the full quantum theory. To first approximation, since the kink is heavy, mesons are reflected or transmitted with some probabilities, while the momentum of the kink is unchanged. But when we work to higher orders, then of course the kink will move as a result. This is all very-complicated.

For more details, see Rajaraman's \emph{Solitons and Instantons}, or Weinberg's \emph{Classical Solutions in Quantum Field Theory}.

The thing that is really hard to understand in the quantum field theory is kink-antikink pair production. This happens when the mesons are very fast, and the theory is highly relativistic. What we have done so far is perturbative and makes the non-relativistic approximation to get the adiabatic picture. It is \emph{very} difficult to understand this.

\subsection{Sine-Gordon kinks}
We end the section by briefly talking about kinks in a different theory, namely the \term{sine-Gordon theory}. In this theory, kinks are often known as \term{soltions} instead.

The sine-Gordon theory is given by the potential
\[
  U(\phi) = 1 - \sin \phi.
\]
Again, we suppress coupling constants, but it is possible to add them back.

The potential looks like
\begin{center}
  \begin{tikzpicture}
    \draw [->](-3, 0) -- (3, 0) node [right] {$\phi$};
    \draw [->] (0, -0.5) -- (0, 2) node [above] {$U(\phi)$};

    \draw [mblue, thick] (0, 0) cos (0.3, 0.7) sin (0.6, 1.4) cos (0.9, 0.7) sin (1.2, 0) cos (1.5, 0.7) sin (1.8, 1.4) cos (2.1, 0.7) sin (2.4, 0) cos (2.7, 0.7) sin (3, 1.4);;
    \draw [mblue, thick, xscale=-1] (0, 0) cos (0.3, 0.7) sin (0.6, 1.4) cos (0.9, 0.7) sin (1.2, 0) cos (1.5, 0.7) sin (1.8, 1.4) cos (2.1, 0.7) sin (2.4, 0) cos (2.7, 0.7) sin (3, 1.4);;

    \node at (1.2, 0) [below] {$2\pi$};
    \node at (2.4, 0) [below] {$4\pi$};
    \node at (-1.2, 0) [below] {$2\pi$};
    \node at (-2.4, 0) [below] {$4\pi$};
  \end{tikzpicture}
\end{center}
Now there are \emph{infinitely many} distinct vacua. In this case, we find we need to pick $W$ such that
\[
  \frac{\d W}{\d \phi} = 2 \sin \frac{1}{2}\phi. % this is dW/dphi instead
\]

\subsubsection*{Static sine-Gordon kinks}
To find the static kinks in the sine-Gordon theory, we again look at the Bogomolny equation. We have to solve
\[
  \frac{\d \phi}{\d x} = 2 \sin \frac{1}{2}\phi.
\]
This can be solved. This involves integrating a $\csc$, and ultimately gives us a solution of
\[
  \phi(x) = 4 \tan^{-1} e^{x - a}.
\]
We can check that this solution interpolates between $0$ and $2\pi$.

\begin{center}
  \begin{tikzpicture}
    \draw [->] (-4, 0) -- (4, 0) node [right] {$x$};
    \draw [->] (0, -0.5) -- (0, 2.5) node [above] {$\phi$};

    \draw [mblue, thick] plot [smooth] coordinates {(-4.0,0.00000) (-3.9,0.00000) (-3.8,0.00000) (-3.7,0.00001) (-3.6,0.00001) (-3.5,0.00001) (-3.4,0.00001) (-3.3,0.00002) (-3.2,0.00002) (-3.1,0.00003) (-3.0,0.00005) (-2.9,0.00006) (-2.8,0.00008) (-2.7,0.00011) (-2.6,0.00015) (-2.5,0.00020) (-2.4,0.00027) (-2.3,0.00037) (-2.2,0.00050) (-2.1,0.00068) (-2.0,0.00091) (-1.9,0.00123) (-1.8,0.00166) (-1.7,0.00224) (-1.6,0.00303) (-1.5,0.00409) (-1.4,0.00552) (-1.3,0.00745) (-1.2,0.01005) (-1.1,0.01357) (-1.0,0.01831) (-0.9,0.02472) (-0.8,0.03336) (-0.7,0.04502) (-0.6,0.06074) (-0.5,0.08190) (-0.4,0.11035) (-0.3,0.14847) (-0.2,0.19922) (-0.1,0.26607) (0.0,0.35251) (0.1,0.46091) (0.2,0.59053) (0.3,0.73548) (0.4,0.88474) (0.5,1.02559) (0.6,1.14850) (0.7,1.24946) (0.8,1.32902) (0.9,1.39011) (1.0,1.43628) (1.1,1.47087) (1.2,1.49666) (1.3,1.51583) (1.4,1.53006) (1.5,1.54061) (1.6,1.54843) (1.7,1.55423) (1.8,1.55852) (1.9,1.56170) (2.0,1.56406) (2.1,1.56580) (2.2,1.56710) (2.3,1.56806) (2.4,1.56877) (2.5,1.56929) (2.6,1.56968) (2.7,1.56997) (2.8,1.57019) (2.9,1.57034) (3.0,1.57046) (3.1,1.57055) (3.2,1.57061) (3.3,1.57066) (3.4,1.57070) (3.5,1.57072) (3.6,1.57074) (3.7,1.57076) (3.8,1.57077) (3.9,1.57077) (4.0,1.57078)};
    % map (\x -> (showFFloat (Just 1) x "", showFFloat (Just 5) (atan ( exp (3*x - 1))) "")) [-4,-3.9..4]

    \draw [dashed] (-4, 1.57078) -- (4, 1.57078);

    \node [left] at (-4, 0) {$0$};
    \node [right] at (4, 1.57078) {$2\pi$};
    \node [circ] at (0.33, 0.785398) {};
    \draw [dashed] (0.33, 0.785398) -- (0.33, 0) node [below] {$a$};
  \end{tikzpicture}
\end{center}

Unlike the $\phi^4$ theory, dynamical multi-kink solutions can be derived \emph{exactly}. Ultimately, this is due to the sine-Gordon equation being \emph{integrable}. For more details, refer to the IID Integrable Systems course. There are multiple ways to obtain these multi-soliton solutions, such as via the B\"acklund transforms, but we will not go into these. Instead, let's just do some examples.

Dynamical multi-kink solutions can be derived exactly. One of the earlier ways to do so was via B\"acklund transforms, but that was very complicated. People later invented better methods, but they are still not very straightforward.

\begin{eg}
  There is a two-kink solution
  \[
    \phi (x, t) = 4 \tan^{-1} \left(\frac{v \sinh \gamma x}{ \cosh \gamma vt}\right),
  \]
  where, as usual, we have
  \[
    \gamma = (1 - v^2)^{-1/2}.
  \]
  For $v = 0.01$, this looks like
  \begin{center}
    \begin{tikzpicture}
      \draw [->] (-4, 0) -- (4, 0) node [right] {$x$};
      \draw [->] (0, -2.5) -- (0, 2.5) node [above] {$\phi$};

      \draw [mblue, thick] plot [smooth] coordinates {(-4.0,-1.56957) (-3.9,-1.56914) (-3.8,-1.56856) (-3.7,-1.56778) (-3.6,-1.56672) (-3.5,-1.56529) (-3.4,-1.56337) (-3.3,-1.56077) (-3.2,-1.55726) (-3.1,-1.55252) (-3.0,-1.54613) (-2.9,-1.53751) (-2.8,-1.52587) (-2.7,-1.51019) (-2.6,-1.48906) (-2.5,-1.46067) (-2.4,-1.42263) (-2.3,-1.37197) (-2.2,-1.30524) (-2.1,-1.21893) (-2.0,-1.11067) (-1.9,-0.98117) (-1.8,-0.83628) (-1.7,-0.68699) (-1.6,-0.54603) (-1.5,-0.42296) (-1.4,-0.32183) (-1.3,-0.24211) (-1.2,-0.18089) (-1.1,-0.13458) (-1.0,-0.09986) (-0.9,-0.07394) (-0.8,-0.05461) (-0.7,-0.04020) (-0.6,-0.02942) (-0.5,-0.02129) (-0.4,-0.01509) (-0.3,-0.01027) (-0.2,-0.00637) (-0.1,-0.00305) (0.0,0.00000) (0.1,0.00305) (0.2,0.00637) (0.3,0.01027) (0.4,0.01509) (0.5,0.02129) (0.6,0.02942) (0.7,0.04020) (0.8,0.05461) (0.9,0.07394) (1.0,0.09986) (1.1,0.13458) (1.2,0.18089) (1.3,0.24211) (1.4,0.32183) (1.5,0.42296) (1.6,0.54603) (1.7,0.68699) (1.8,0.83628) (1.9,0.98117) (2.0,1.11067) (2.1,1.21893) (2.2,1.30524) (2.3,1.37197) (2.4,1.42263) (2.5,1.46067) (2.6,1.48906) (2.7,1.51019) (2.8,1.52587) (2.9,1.53751) (3.0,1.54613) (3.1,1.55252) (3.2,1.55726) (3.3,1.56077) (3.4,1.56337) (3.5,1.56529) (3.6,1.56672) (3.7,1.56778) (3.8,1.56856) (3.9,1.56914) (4.0,1.56957)};
      \draw [mred, thick] plot [smooth] coordinates {(-4.0,-1.53726) (-3.9,-1.52555) (-3.8,-1.50975) (-3.7,-1.48847) (-3.6,-1.45988) (-3.5,-1.42157) (-3.4,-1.37057) (-3.3,-1.30340) (-3.2,-1.21659) (-3.1,-1.10779) (-3.0,-0.97784) (-2.9,-0.83269) (-2.8,-0.68347) (-2.7,-0.54286) (-2.6,-0.42031) (-2.5,-0.31974) (-2.4,-0.24053) (-2.3,-0.17975) (-2.2,-0.13381) (-2.1,-0.09939) (-2.0,-0.07374) (-1.9,-0.05467) (-1.8,-0.04052) (-1.7,-0.03002) (-1.6,-0.02224) (-1.5,-0.01648) (-1.4,-0.01221) (-1.3,-0.00904) (-1.2,-0.00670) (-1.1,-0.00496) (-1.0,-0.00367) (-0.9,-0.00271) (-0.8,-0.00200) (-0.7,-0.00147) (-0.6,-0.00108) (-0.5,-0.00078) (-0.4,-0.00055) (-0.3,-0.00038) (-0.2,-0.00023) (-0.1,-0.00011) (0.0,0.00000) (0.1,0.00011) (0.2,0.00023) (0.3,0.00038) (0.4,0.00055) (0.5,0.00078) (0.6,0.00108) (0.7,0.00147) (0.8,0.00200) (0.9,0.00271) (1.0,0.00367) (1.1,0.00496) (1.2,0.00670) (1.3,0.00904) (1.4,0.01221) (1.5,0.01648) (1.6,0.02224) (1.7,0.03002) (1.8,0.04052) (1.9,0.05467) (2.0,0.07374) (2.1,0.09939) (2.2,0.13381) (2.3,0.17975) (2.4,0.24053) (2.5,0.31974) (2.6,0.42031) (2.7,0.54286) (2.8,0.68347) (2.9,0.83269) (3.0,0.97784) (3.1,1.10779) (3.2,1.21659) (3.3,1.30340) (3.4,1.37057) (3.5,1.42157) (3.6,1.45988) (3.7,1.48847) (3.8,1.50975) (3.9,1.52555) (4.0,1.53726)};
      % let v = 0.01 in let g = sqrt (1 / (1 - v * v)) in let t = 400 in map (\x -> (showFFloat (Just 1) x "", showFFloat (Just 5) (atan (v * sinh (g * x * 3) / cosh (v * g * t))) "")) [-4,-3.9..4]

      \draw [dashed] (-4, 1.57078) -- (4, 1.57078);
      \draw [dashed] (-4, -1.57078) -- (4, -1.57078);

      \node [right] at (4, 1.57078) {$2\pi$};
      \node [left] at (-4, -1.57078) {$-2\pi$};

      \node [mblue, left] at (1.73, 0.785) {\small$t = 0$};
      \node [mred, right] at (2.85, 0.785) {\small$t = 400$};
    \end{tikzpicture}
  \end{center}

  Note that since $\phi(x, t) = \phi(x, -t)$, we see that this solution involves two solitons at first approaching each other, and then later bouncing off. Thus, the two kinks \emph{repel} each other. This is not too surprising, because when we did kinks in $\phi^4$ theory, we saw that a kink and an anti-kink repelled.

  We can again compute the force just like the $\phi^4$ theory, but alternatively, since we have a full, exact solution, we can work it out directly from the solution! The answers, fortunately, agree. If we do the computations, we find that the point of closest approach is $\sim 2 \log \left(\frac{2}{v}\right)$ if $v$ is small.
\end{eg}

There are some important comments to make. In the sine-Gordon theory, we can have very complicated interactions between kinks and anti-kinks, and these can connect vastly different vacua. However, \emph{static} solutions must join $2n\pi$ and $2(n \pm 1)\pi$ for some $n$, because if we want to join vacua further apart, we will have more than one kink, and they necessarily interact.

If we have multiple kinks and anti-kinks, then teach of these things can have their own velocity, and we might expect some very complicated interaction between them, such as annihilation and pair production. But remarkably, the interactions are \emph{not} complicated. If we try to do numerical simulations, or use the exact solutions we find, we see that we do not have energy loss due to ``radiation''. Instead, the solitons remain very well-structured solitons. This, again, is due to the theory being integrable.

\subsubsection*{Topology of the sine-Gordon equation}
There are also a lot of interesting things we can talk about without going into details about what the solutions look like.

The important realization is that our potential is periodic in $\phi$. For the sine-Gordon theory, it is much better to think of this as a field modulo $2\pi$, i.e.\ as a function
\[
  \phi: \R \to S^1.
\]
In this language, the boundary condition is that $\phi(x) = 0 \bmod 2\pi$ as $x \to \pm \infty$. Thus, instead of thinking of the kink as joining two vacua, we can think of it as ``winding around the circle'' instead.

We can go further. Since the boundary conditions of $\phi$ are the same on two sides, we can join the ends of the domain together, and we can think of $\phi$ as a map
\[
  \phi: S^1 \to S^1
\]
instead. This is a \term{compactification} of the space.

Topologically, such maps are classified by their \term{winding number}, or the \term{degree}, which we denote \term{$Q$}. This is a topological (homotopy) invariant of a map, and is preserved under continuous deformations of the field. Thus, it is preserved under time evolution of the field.

Intuitively, the winding number is just how many times we go around the circle. There are multiple (equivalent) ways of making this precise.

\separator

The first way, which is the naive way, is purely topological. We simply have to go back to the first picture, where we regard $\phi$ as a real value. Suppose the boundary values are
\[
  \phi(-\infty) = 2 n_L \pi,\quad \phi(\infty) = 2 n_R \pi.
\]
Then we set the winding number to be $Q = n_R - n_L$.

Topologically, we are using the fact that $\R$ is the \term{universal covering space}\index{covering space} of the circle, and thus we are really looking at the induced map on the fundamental group of the circle.

\begin{eg}
  As we saw, a single kink has $Q = 1$.
  \begin{center}
    \begin{tikzpicture}
      \draw [->] (-4, 0) -- (4, 0) node [right] {$x$};
      \draw [->] (0, -0.5) -- (0, 2.5) node [above] {$\phi$};

      \draw [mblue, thick] plot [smooth] coordinates {(-4.0,0.00000) (-3.9,0.00000) (-3.8,0.00000) (-3.7,0.00001) (-3.6,0.00001) (-3.5,0.00001) (-3.4,0.00001) (-3.3,0.00002) (-3.2,0.00002) (-3.1,0.00003) (-3.0,0.00005) (-2.9,0.00006) (-2.8,0.00008) (-2.7,0.00011) (-2.6,0.00015) (-2.5,0.00020) (-2.4,0.00027) (-2.3,0.00037) (-2.2,0.00050) (-2.1,0.00068) (-2.0,0.00091) (-1.9,0.00123) (-1.8,0.00166) (-1.7,0.00224) (-1.6,0.00303) (-1.5,0.00409) (-1.4,0.00552) (-1.3,0.00745) (-1.2,0.01005) (-1.1,0.01357) (-1.0,0.01831) (-0.9,0.02472) (-0.8,0.03336) (-0.7,0.04502) (-0.6,0.06074) (-0.5,0.08190) (-0.4,0.11035) (-0.3,0.14847) (-0.2,0.19922) (-0.1,0.26607) (0.0,0.35251) (0.1,0.46091) (0.2,0.59053) (0.3,0.73548) (0.4,0.88474) (0.5,1.02559) (0.6,1.14850) (0.7,1.24946) (0.8,1.32902) (0.9,1.39011) (1.0,1.43628) (1.1,1.47087) (1.2,1.49666) (1.3,1.51583) (1.4,1.53006) (1.5,1.54061) (1.6,1.54843) (1.7,1.55423) (1.8,1.55852) (1.9,1.56170) (2.0,1.56406) (2.1,1.56580) (2.2,1.56710) (2.3,1.56806) (2.4,1.56877) (2.5,1.56929) (2.6,1.56968) (2.7,1.56997) (2.8,1.57019) (2.9,1.57034) (3.0,1.57046) (3.1,1.57055) (3.2,1.57061) (3.3,1.57066) (3.4,1.57070) (3.5,1.57072) (3.6,1.57074) (3.7,1.57076) (3.8,1.57077) (3.9,1.57077) (4.0,1.57078)};
    % map (\x -> (showFFloat (Just 1) x "", showFFloat (Just 5) (atan ( exp (3*x - 1))) "")) [-4,-3.9..4]

      \draw [dashed] (-4, 1.57078) -- (4, 1.57078);

      \node [left] at (-4, 0) {$0$};
      \node [right] at (4, 1.57078) {$2\pi$};
    \end{tikzpicture}
  \end{center}
\end{eg}

Thus, we can think of the $Q$ as the \emph{net soliton number}.

\separator

But this construction we presented is rather specific to maps from $S^1$ to $S^1$. We want something more general that can be used for more complicated systems. We can do this in a more ``physics'' way. We note that there is a \emph{topological} current
\[
  j^\mu = \frac{1}{2\pi} \varepsilon^{\mu\nu} \partial_\nu \phi,
\]
where $\varepsilon^{\mu\nu}$ is the anti-symmetric tensor in $1 + 1$ dimensions, chosen so that $\varepsilon^{01} = 1$.

In components, this is just
\[
  j^\mu = \frac{1}{2\pi} (\partial_x \phi, - \partial_t \phi).
\]
This is conserved because of the symmetry of mixed partial derivatives, so that
\[
  \partial_\mu j^\mu = \frac{1}{2\pi} \varepsilon^{\mu\nu} \partial_\mu \partial_\nu \phi = 0.
\]
As usual, a current induces a conserved charge
\[
  Q = \int j^0 \;\d x = \frac{1}{2\pi} \int \partial_x \phi \;\d x = \frac{1}{2\pi} (\phi(\infty) - \phi(-\infty)) = n_R - n_L,
\]
which is the formula we had earlier.

Note that all these properties do not depend on $\phi$ satisfying any field equations! It is completely topological.

\separator

Finally, there is also a differential geometry way of defining this. We note that the target space $S^1$ has a normalized volume form $\omega$ so that
\[
  \int_{S^1} \omega = 1.
\]
For example, we can take
\[
  \omega = \frac{1}{2\pi}\;\d \phi.
\]
Now, given a mapping $\phi: \R \to S^1$, we can pull back the volume form to obtain
\[
  \phi^* \omega = \frac{1}{2\pi} \frac{\d \phi}{\d x} \;\d x.
\]
We can then define the degree of the map to be
\[
  Q = \int \phi^* \omega = \frac{1}{2\pi}\int_{-\infty}^\infty \frac{\d \phi}{\d x}\;\d x.
\]
This is exactly the same as the formula we obtained using the current!

Note that even though the volume form is normalized on $S^1$, the integral when pulled back is not $1$. We can imagine this as saying if we wind around the circle $n$ times, then after pulling back, we would have pulled back $n$ ``copies'' of the volume form, and so the integral will be $n$ times that of the integral on $S^1$.

\separator

We saw that these three definitions gave the same result, and different definitions have different benefits. For example, in the last two formulations, it is not \emph{a priori} clear that the winding number has to be an integer, while this is clear in the first formulation.

\section{Vortices}
We are now going to start studying \emph{vortices}. These are topological solitons in two space dimensions. While we mostly studied $\phi^4$ kinks last time, what we are going to do here is more similar to the sine-Gordon theory than the $\phi^4$ theory, as it is largely topological in nature.

A lot of the computations we perform in the section are much cleaner when presented using the language of differential forms. However, we shall try our best to provide alternative versions in coordinates for the easily terrified.

\subsection{Topological background}
\subsubsection*{Sine-Gordon kinks}
We now review what we just did for sine-Gordon kinks, and then try to develop some analogous ideas in higher dimension. The sine-Gordon equation is given by
\[
  \frac{\partial^2 \theta}{\partial t^2} - \frac{\partial^2 \theta}{\partial x^2} + \sin \theta = 0.
\]

We want to choose boundary conditions so that the energy has a chance to be finite. The first part is, of course, to figure out what the energy is. The energy-momentum conservation equation given by Noether's theorem is
\[
  \partial_t \left(\frac{\theta_t^2 + \theta_x^2}{2} + (1 - \cos \theta)\right) + \partial_x (-\theta_t \theta_x) = \partial_\mu P^\mu = 0.
\]
The energy we will be considering is thus
\[
  E = \int_{\R} P^0 \;\d x = \int_\R \left(\frac{\theta_t^2 + \theta_x^2}{2} + (1 - \cos \theta)\right)\;\d x.
\]
Thus, to obtain finite energy, we will want $\theta(x) \to 2n_{\pm} \pi$ for some integers $n_{\pm}$ as $x \to \pm \infty$. What is the significance of this $n_{\pm}$?
\begin{eg}
  Consider the basic kink
  \[
    \theta_K(x) = 1 + \tan^{-1} e^x.
  \]
  Picking the standard branch of $\tan^{-1}$, this kink looks like
  \begin{center}
    \begin{tikzpicture}
      \draw [->] (-4, 0) -- (4, 0) node [right] {$x$};
      \draw [->] (0, -0.5) -- (0, 2.5) node [above] {$\phi$};

      \draw [mblue, thick] plot [smooth] coordinates {(-4.0,0.00001) (-3.9,0.00001) (-3.8,0.00001) (-3.7,0.00002) (-3.6,0.00002) (-3.5,0.00003) (-3.4,0.00004) (-3.3,0.00005) (-3.2,0.00007) (-3.1,0.00009) (-3.0,0.00012) (-2.9,0.00017) (-2.8,0.00022) (-2.7,0.00030) (-2.6,0.00041) (-2.5,0.00055) (-2.4,0.00075) (-2.3,0.00101) (-2.2,0.00136) (-2.1,0.00184) (-2.0,0.00248) (-1.9,0.00335) (-1.8,0.00452) (-1.7,0.00610) (-1.6,0.00823) (-1.5,0.01111) (-1.4,0.01499) (-1.3,0.02024) (-1.2,0.02732) (-1.1,0.03687) (-1.0,0.04975) (-0.9,0.06710) (-0.8,0.09047) (-0.7,0.12185) (-0.6,0.16382) (-0.5,0.21953) (-0.4,0.29255) (-0.3,0.38616) (-0.2,0.50193) (-0.1,0.63760) (0.0,0.78540) (0.1,0.93320) (0.2,1.06887) (0.3,1.18464) (0.4,1.27824) (0.5,1.35126) (0.6,1.40698) (0.7,1.44895) (0.8,1.48033) (0.9,1.50369) (1.0,1.52105) (1.1,1.53393) (1.2,1.54348) (1.3,1.55056) (1.4,1.55580) (1.5,1.55969) (1.6,1.56257) (1.7,1.56470) (1.8,1.56628) (1.9,1.56745) (2.0,1.56832) (2.1,1.56896) (2.2,1.56944) (2.3,1.56979) (2.4,1.57005) (2.5,1.57024) (2.6,1.57039) (2.7,1.57049) (2.8,1.57057) (2.9,1.57063) (3.0,1.57067) (3.1,1.57070) (3.2,1.57073) (3.3,1.57075) (3.4,1.57076) (3.5,1.57077) (3.6,1.57078) (3.7,1.57078) (3.8,1.57079) (3.9,1.57079) (4.0,1.57079)};
    % map (\x -> (showFFloat (Just 1) x "", showFFloat (Just 5) (atan ( exp (3*x - 1))) "")) [-4,-3.9..4]

      \draw [dashed] (-4, 1.57078) -- (4, 1.57078);

      \node [left] at (-4, 0) {$0$};
      \node [right] at (4, 1.57078) {$2\pi$};
    \end{tikzpicture}
  \end{center}
  This goes from $\theta = 0$ to $\theta = 2\pi$.
\end{eg}

To better understand this, we can think of $\theta$ as an angular variable, i.e.\ we identify $\theta \sim \theta + 2k\pi$ for any $k \in \Z$. This is a sensible thing because the energy density and the equation etc.\ are unchanged when we shift everything by $2k\pi$. Thus, $\theta$ is not taking values in $\R$, but in $\R/2\pi\Z \cong S^1$.

Thus, for each fixed $t$, our field $\theta$ is a map
\[
  \theta: \R \to S^1.
\]
The number $Q = n_+ - n_-$ equals the number of times $\theta$ covers the circle $S^1$ on going from $x = -\infty$ to $x = +\infty$. This is the winding number, which is interpreted as the topological charge.

As we previously discussed, we can express this topological charge as the integral of some current. We can write
\[
  Q = \frac{1}{2\pi} \int_{\theta(\R)} \d \theta = \frac{1}{2\pi} \int_{-\infty}^\infty \frac{\d \theta}{\d x}\;\d x.
\]
Note that this formula automatically takes into account the orientation. This is the form that will lead to generalization in higher dimensions.

This function $\frac{\d \theta}{\d x}$ appearing in the integral has the interpretation as a topological charge. Note that there is a topological conservation law
\[
  \partial_\mu j^\mu = \frac{\partial j^0}{\partial t} + \frac{\partial j^1}{\partial x} = 0,
\]
where
\[
  j^0 = \theta_x,\quad j^1 = -\theta_t.
\]
This conservation law is not a consequence of the field equations, but merely a mathematical identity, namely the commutation of partial derivatives.

\subsubsection*{Two dimensions}
In the sine-Gordon kink, the target space was a circle $S^1$. This time, we are concerned with the \term{unit disk}\index{$D$}
\[
  D = \{(x^1, x^2): |\mathbf{x}|^2 < 1\} \subseteq \R^2.
\]
We will then consider fields
\[
  \Phi: D \to D.
\]
In the case of a sine-Gordon kink, we still cared about moving solitons. However, here we will mostly work with static solutions, and study fields at a fixed time. Thus, there is no time variable appearing.

Using the canonical isomorphism $\R^2 \cong \C$, we can think of the target space as the unit disk in the complex plane, and write the field as
\[
  \Phi = \Phi^1 + i \Phi^2.
\]
However, we will usually view the $D$ in the domain as a real space instead.

We will impose some boundary conditions. We pick any function $\chi: S^1 \to \R$, and consider
\[
  g = e^{i\chi}: S^1 \to S^1 \subseteq D.
\]
We then require
\[
  \Phi_{\partial D} = g.
\]
In particular, $\Phi$ must send the boundary to the boundary.

Now the target space $D$ has a canonical choice of measure $\d x^1 \wedge \d x^2$. Then we can expect the new topological charge to be given by
\[
  Q = \frac{1}{\pi} \int_{\Phi(D)} \d \Phi^1 \wedge \d \Phi^2 = \frac{1}{\pi} \int_D \det
  \begin{pmatrix}
    \frac{\partial \Phi^1}{\partial x^1} & \frac{\partial \Phi^1}{\partial x^2}\\
    \frac{\partial \Phi^2}{\partial x^1} & \frac{\partial \Phi^2}{\partial x^2}
  \end{pmatrix}\;\d x^1 \wedge \d x^2.
\]
Thus, the charge density is given by
\[
  j^0 = \frac{1}{2} \varepsilon_{ab} \varepsilon_{ij} \frac{\partial \Phi^a}{\partial x^i} \frac{\partial \Phi^b}{\partial x^j}.
\]
Crucially, it turns out this charge density is a total derivative, i.e.\ we have
\[
  j^0 = \frac{\partial V^i}{\partial x^i}
\]
for some function $V$. It is not immediately obvious this is the case. However, we can in fact pick
\[
  V^i = \frac{1}{2} \varepsilon_{ab}\varepsilon^{ij} \Phi^a \frac{\partial \Phi^b}{\partial x^j}.
\]
To see this actually works, we need to use the anti-symmetry of $\varepsilon^{ij}$ and observe that
\[
  \varepsilon^{ij} \frac{\partial^2 \Phi^b}{\partial x^i \partial x^j} = 0
\]
Equivalently, using the language of differential forms, we view the current $j^0$ as the $2$-form
\[
  j^0 = \d \Phi^1 \wedge \d \Phi^2 = \d (\Phi^1\;\d \Phi^2) = \frac{1}{2} \d (\Phi^1 \;\d \Phi^2 - \Phi^2 \;\d \Phi^1).
\]
By the divergence theorem, we find that
\[
  Q = \frac{1}{2\pi}\oint_{\partial D} \Phi^1 \;\d \Phi^2 - \Phi^2 \;\d \Phi^1.
\]
We then use that on the boundary,
\[
  \Phi^1 = \cos \chi,\quad \Phi^2 = \sin \chi.
\]
Then we can write this as
\[
  Q = \frac{1}{2\pi} \oint_{\partial D} (\cos^2 \chi + \sin^2 \chi) \;\d \chi = \frac{1}{2\pi} \oint_D \;\d \chi = N.
\]
Thus, the charge is just the winding number of $g$!

%But what actually is the degree telling us? Let's fix a boundary condition $\Phi|_{\partial D} = g = e^{i\chi}$, where $\chi: \partial D \cong S^1 \to \R$ is a real function. Since $g$ is now a map from the boundary circle to the boundary circle. So $g$ itself has got a winding number,
%\[
%  \frac{1}{2\pi} \int_0^{2\pi} \frac{\d \chi}{\d \theta} \;\d \theta.
%\]
%Let's relate the degree of $\Phi$ to the winding number of $g$. To do so, we will need the fact that the Jacobian determinant is a total derivative, so that we can apply Green's identity. Thus, we want to write $j^0$ as
%\[
%  j^0 = \frac{\partial V^i}{\partial x^i}
%\]
%for some $V^i$. It might not be immediately obvious how we can pick such a $V^i$, but notice that
%\[
%  \varepsilon_{ij} \frac{\partial^2 \Phi^b}{\partial x^i \partial x^j} = 0
%\]
%by anti-symmetry. So we can in fact pick
%\[
%  V^i = \frac{1}{2} \varepsilon_{ab}\varepsilon_{ij} \Phi^a \frac{\partial \Phi^b}{\partial x^j}.
%\]
%Then in this case, we have
%\[
%  \int_D j^0 \;\d x = \int \left(\frac{\partial V^1}{\partial x^1} + \frac{\partial V^2}{\partial x^2}\right)\;\d x^1 \;\d x^2 = \oint_{\partial D} V^1 \;\d x^2 - V^2 \;\d x^1.
%\]
%On the boundary, we know $\Phi^1 = \cos \chi$ and $\Phi^2 = \sin \chi$. Substituting that in, we can calculate $V^1$ and $V^2$ on the boundary, and then work out the integral to be
%\[
%  \int_D j\;\d x = \frac{1}{2} \oint_D \;\d \chi = \pi N,
%\]
%where $N$ is the winding number of $g: S^1 \to S^1$.
%
%This is a general phenomenon. Often, the appearance of anti-symmetric combinations of derivatives will allow us to reduce complicated quantities into total derivatives. Then this will allow us to relate things that \emph{a priori} depends on the value of $\Phi$ everywhere into something that depends only on the boundary values.
%The next interesting thing to notice is that the charge $Q$ is invariant under scaling of the domain. If we scale $x$ by $R$, then $\d x^1 \wedge \d x^2$ gets scaled up by $R^2$, but the Jacobian gets scaled down by $R^2$. Thus, an interesting thing to do is to take take $R \to \infty$. Then the domain becomes all of $\R^2$.

Now notice that our derivation didn't really depend on our domain being $D$. It could have been any region bounded by a simple closed curve in $\R^2$. In particular, we can take it to be a disk of arbitrary radius.

What we are \emph{actually} interested in is a field
\[
  \Phi: \R^2 \to D.
\]
We then impose asymptotic boundary conditions
\[
  \Phi \sim g = e^{i\chi}
\]
as $|x| \to \infty$. We can still define the degree by
\[
  Q = \frac{1}{\pi} \int_{\R^2} j^0 \;\d x^1 \;\d x^2 = \frac{1}{\pi} \lim_{R \to \infty}\int_{RD} j^0 \;\d x^1 \;\d x^2
\]
This is then again the winding number of $g$.
%So our field is now a mapping $\Phi: \R^2 \to D$. We then set the asymptotic boundary condition $\Phi \sim g = e^{i \chi}$ as $|x| \to \infty$. For these maps, we have the same formula for degree, namely
%\[
%  Q = \frac{1}{\pi} \int_{\R^2} \frac{1}{2} \varepsilon_{ab} \varepsilon_{ij} \partial_i \Phi^a \partial_j \Phi^b \;\d x^1 \;\d x^2 = \frac{1}{2\pi} \lim_{R\to \infty} \oint_{|x| = R} \d \chi \in \Z.
%\]
%This is going to be our basic topological quantity for the vortices. Its existence, and being non-zero, is what gives rise to a stable vortex.

It is convenient to rewrite this in terms an inner product. $\R^2$ itself has an inner product, and if under the identification $\R^2 \cong \C$, the inner product can be written as
\[
  (a, b) = \frac{\bar{a} b + a \bar{b}}{2}.
\]
Usually, this expression isn't that useful, and we will just use the fact that for real numbers $a$ and $b$, we have
\[
  (a, b) = (ai, bi) =  ab, \quad (ai, b) = (a, bi) = 0.
\]
In particular, we can evaluate
\[
  (i \Phi, \d \Phi) = (i \Phi^1 - \Phi^2, \d \Phi^1 + i \d \Phi^2) = \Phi^1\;\d \Phi^2 - \Phi^2\;\d \Phi^1.
\]
This is just (twice) the $V$ we found! So we can write our charge as
\[
  Q = \frac{1}{2\pi} \lim_{R \to \infty} \oint_{|x| = R} (i \Phi, \d \Phi).
\]
Often, we will refer to $(i \Phi, \d \Phi)$ itself as the current, instead of $j^0 = \frac{1}{2} \d (i \Phi, \d \Phi)$.

%We can rewrite this in terms of complex numbers. Instead of viewing a $\Phi$ as a function valued in $D \subseteq \R^2$, we view it as a function $\Phi: D \to \C$. We already implicitly did so when we wrote the boundary condition as $e^{i \chi}$. In terms of complex numbers, we can write the standard inner product on $\R^2$ as
%\[
%  (a, b) = \frac{\bar{a} b + a \bar{b}}{2}.
%\]
%Then we can write the integral for the charge as
%\[
%  \oint_{\partial D}\;\d \chi = \oint_{\partial D} (i\Phi, \d \Phi).
%\]
%This formulation is sometimes more convenient for calculations.

We have actually seen this current before. In quantum mechanics, we have the Schr\"odinger's equation
\[
  i \frac{\partial \Phi}{\partial t} = - \frac{1}{2m} \Delta \Phi + V(x) \Phi,\quad \Delta = \nabla^2.
\]
We saw that probability is conserved in such a theory, and the probability conservation law is given by
\[
  \frac{1}{2} \partial_t (\Phi, \Phi) + \frac{1}{2m} \nabla \cdot (i \Phi, \nabla \Phi) = 0.
\]
What appears here is just our topological current!

%The final comment, which we will not actually use, is that if we have a holomorphic function $f$, then consider the integral
%\[
% \frac{1}{2\pi i} \oint_C \frac{f'(z)}{f(z) - w}\;\d z.
%\]
%Using the residue theorem, we can see that this is the number of times $f$ takes value $w$, counted with multiplicity, within the area bounded by $C$. This is the \term{local degree} of the mapping. Indeed,
%\[
% \Res\left(\frac{f'(z)}{f(z) - w}, z_i\right) = n_i,
%\]
%where $f(z) - w = (z - z_i)^{n_i'} h(z)$, with $h(z_i) \not= 0$.
%
%But we can think about this in a different way. The integrand is an exact differential
%\[
% \frac{1}{2\pi i} [\log (f(z) - w)],
%\]
%and if we think about this carefully, this the change in argument of $f(z) - w$ around the curve.

\subsection{Global \texorpdfstring{$U(1)$}{U(1)} Ginzburg--Landau vortices}
We now put the theory into use. We are going to study \emph{Ginzburg--Landau vortices}\term{Ginzburg--Landau vortex}. Our previous discussion involved a function taking values in the unit disk. We will not impose such a restriction on our vortices. However, we will later see that any solution must take values in $D$.

The potential of the Ginzburg--Landau theory is given by
\[
  V(\Phi) = \frac{1}{2} \int_{\R^2} \left((\nabla \Phi, \nabla \Phi) + \frac{\lambda}{4} (1 - (\Phi, \Phi))^2 \right)\;\d x^1 \;\d x^2.
\]
where $\lambda > 0$ is some constant.

Note that the inner product is invariant under phase rotation, i.e.
\[
  (e^{i\chi} a, e^{i\chi b}) = (a, b)
\]
for $\chi \in \R$. So in particular, the potential satisfies
\[
  V(e^{i\chi}\Phi) = V(\Phi).
\]
Thus, our theory has a \term{global $\U(1)$ symmetry}\index{$\U(1)$ symmetry!global}.

The Euler--Lagrange equations of this theory says
\[
  - \Delta \Phi = \frac{\lambda}{2}(1 - |\Phi|^2) \Phi.
\]
This is the \term{ungauged Ginzburg--Landau} \emph{equation}\index{Ginzburg--Landau!ungauged}.

To justify the fact that our $\Phi$ takes values in $D$, we use the following lemma:
\begin{lemma}
  At any maximum point of $|\Phi|$, we have $|\Phi| \leq 1$.
\end{lemma}

\begin{proof}
  Consider the function
  \[
    W(x) = 1 - |\Phi(x)|^2.
  \]
  Then we want to show that $W \geq 0$ when $W$ is minimized. We note that if $W$ is at a minima, then the Hessian matrix must have non-negative eigenvalues. So, taking the trace, we must have $\Delta W \geq 0$. Now we can compute $\Delta W$ directly. We have
  \begin{align*}
    \nabla W &= -2 (\Phi, \nabla \Phi)\\
    \Delta W &= \nabla^2 W \\
    &= - 2(\Phi, \Delta \Phi) - 2(\nabla \Phi, \nabla \Phi)\\
    &= \lambda |\Phi|^2 W - 2 |\nabla \Phi|^2.
  \end{align*}
  Thus, we can rearrange to write this to say
  \[
    2 |\nabla \Phi|^2 + \Delta W = \lambda |\Phi|^2 W.
  \]
  But we know $2 |\nabla \Phi|^2 \geq 0$, and $\Delta W \geq 0$. So we must have $W \geq 0$.
\end{proof}

By itself, this doesn't force $|\Phi| \in [0, 1]$, since we could imagine $\Phi$ having no maximum. However, if we prescribe boundary conditions such that $|\Phi| = 1$ on the boundary, then this would indeed for $|\Phi| \leq 1$ everywhere.

Often, we can think of $\Phi$ as some ``complex order parameter'', and so $|\Phi|$ represents some sort of proportion. Then the condition $|\Phi| \leq 1$ is very natural.

The quantities we are interested in are \emph{vortices}.

\begin{defi}[Gintzburg--Landau vortex]\index{Gintzburg--Landau vortex}\index{vortex}
  A basic \emph{Ginzburg--Landau vortex} of charge $N$ is a (smooth) solution of the form
  \[
    \Phi = f_N(r) e^{iN\theta}
  \]
  in polar coordinates $(r, \theta)$. Moreover, we require that $f_N(r) \to 1$ as $r \to \infty$.
\end{defi}
Note that for $\Phi$ to be a smooth solution, we must have $f_N(0) = 0$. In fact, a bit more analysis shows that we must have $f_N = O(r^N)$ as $r \to 0$. Solutions for $f_N$ do exist, and they look roughly like this:

\begin{center}
  \begin{tikzpicture}
    \draw [->] (-0.5, 0) -- (5, 0) node [right] {$x$};
    \draw [->] (0, -0.5) -- (0, 3) node [above] {$m$};

    \draw [mblue, thick, domain=0:5] plot [smooth] (\x, {2 * \x^4 / (1 + \x^4)});

    \draw [dashed] (-1, 2) -- (5, 2);

  \end{tikzpicture}
\end{center}

In the case of $N = 1$, we can visualize the field $\Phi$ as a vector field on $\C$.. Then it looks like
\begin{center}
  \begin{tikzpicture}
    \draw [->] (-3, 0) -- (3, 0);
    \draw [->] (0, -3) -- (0, 3);

    \foreach \t in {0,30,60,90,120,150,180,210,240,270,300,330,360} {
      \begin{scope}[rotate=\t]
        \foreach \x in {0.5, 1, 1.5, 2, 2.5} {
          \pgfmathsetmacro\arlen{0.5 * (\x^4 / (1 + \x^4))^2}
          \draw [-latex'] (\x, 0) -- +(\arlen, 0);
        }
      \end{scope}
    }
  \end{tikzpicture}
\end{center}
This is known as a \emph{$2$-dimensional hedgehog}.

For general $N$, it might be more instructive to look at how the current looks like. Recall that the current is defined by $j = (i\Phi, \d \Phi)$. We can write this more explicitly as
\begin{align*}
  (i\Phi, \d \Phi) &= (if_N e^{iN\theta}, (\d f_N) e^{iN\theta} + i f_N N\;\d \theta e^{iN\theta})\\
  &= (if_N,\d f_N + i f_N N\;\d \theta).
\end{align*}
We see that $if_N$ and $\d f_N$ are orthogonal, while $i f_N$ and $i f_N N \d \theta$ are parallel. So the final result is
\[
  (i\Phi, \d \Phi) = f_N^2 N\;\d \theta.
\]
So the current just looks like spirals:

\begin{center}
  \begin{tikzpicture}
    \draw [->] (-3, 0) -- (3, 0);
    \draw [->] (0, -3) -- (0, 3);

    \foreach \t in {0,30,60,90,120,150,180,210,240,270,300,330,360} {
      \begin{scope}[rotate=\t]
        \foreach \x in {0.5, 1, 1.5, 2, 2.5} {
          \pgfmathsetmacro\arlen{0.5 * (\x^4 / (1 + \x^4))^2}
          \draw [-latex'] (\x, 0) -- +(0, \arlen);
        }
      \end{scope}
    }
  \end{tikzpicture}
\end{center}

As $|x| \to \infty$, we have $f_N \to 1$. So the winding number is given by as before, and we can compute the winding number of this system to be
\[
  \frac{1}{2\pi} \lim_{R \to \infty} \oint j^0 \;\d \theta = N.
\]
Since the winding number of these systems are discrete quantities, it makes the vortex system stable.

This theory looks good so far. However, it turns out this model has a problem --- the energy is infinite! We can expand out $V(f_N e^{iN \theta})$, and see it is a sum of a few non-negative terms. We will focus on the $\frac{\partial}{\partial \theta}$ term. We obtain
\begin{align*}
  V(f_N e^{iN\theta}) &\geq \int \frac{1}{r^2} \left|\frac{\partial \Phi}{\partial \theta}\right|^2 r\;\d r\;\d \theta\\
  &= N^2 \int \frac{1}{r^2} f_N^2 r\;\d r\;\d \theta \\
  &= 2\pi N \int_0^\infty \frac{f_N^2}{r}\;\d r.
\end{align*}
Since $f_N \to 1$ as $r \to \infty$, we see that the integral diverges logarithmically.

This is bad. We can heuristically understand this by decomposing $\Phi$ into two components. At each fixed point, we can decompose $\Phi$ into a mode parallel to $\Phi$ and a mode perpendicular to $\Phi$. These correspond to the radial and angular modes directly. Moreover, we will see that the parallel mode is massive, while the perpendicular mode is massless.

We saw that the energy divergence is due to the angular part of the action, so we see that it is the massless mode that leads to problems.

We can also see the difference between massless and massive modes in a different setting. Consider the equation
\[
  -\Delta u + M^2 u = f.
\]
Working in three dimensions, the solution is given by
\[
  u(x) = \frac{1}{4\pi} \int \frac{e^{-M|x - y|}}{|x - y|} f(y)\;\d y.
\]
Thus, the Green's function is given by
\[
  G(x) = \frac{e^{-M|x|}}{|x|}.
\]
If the system is massless, i.e.\ $M = 0$, then this decays as $\frac{1}{|x|}$. However, if the system is massive with $M > 0$, then this decays exponentially as $|x| \to \infty$.

So how do we figure out the massive and massless modes? We do not have a genuine decomposition of $\Phi$ into ``parallel'' and ``perpendicular'' modes, because what is parallel and what is perpendicular depends on the current value of $\Phi$.

Thus, to make sense of this, we have to consider small fluctuations around a fixed configuration $\Phi$. We suppose $\Phi$ is a solution to the field equations. Then $\frac{\delta V}{\delta \Phi} = 0$. Thus, for small variations $\Phi \mapsto \Phi + \varepsilon \varphi$, we have
\[
  V(\Phi + \varepsilon \varphi) = V(\Phi) + \varepsilon^2 \int \left(|\nabla \varphi|^2 + \lambda (\varphi, \Phi)^2 - \frac{2\lambda}{4}(1 - |\Phi|^2)|\varphi|^2\right) \;\d x + O(\varepsilon^3).
\]
Ultimately, we are interested in the asymptotic behaviour $|x| \to \infty$, in which case $1 - |\Phi|^2 \to 0$. Moreover, $|\Phi| \to 1$ implies $(\varphi, \Phi)$ becomes a projection into the direction of $\Phi$ Then the potential becomes
\[
  V(\Phi + \varepsilon \varphi) = V(\Phi) + \int \left(|\nabla \varphi|^2 + \lambda |\varphi^{\mathrm{parallel}}|^2\right)\;\d x + O(\varepsilon^3).
\]
Thus, the parallel mode is massive, while the perpendicular mode is massless.

%
%The divergence of energy arises because if we consider fluctuations $\Phi + \varepsilon \varphi$, and look at the energy
%\[
% V(\Phi + \varepsilon \varphi) = V(\Phi) + \varepsilon \int \frac{\delta V}{\delta \Phi} (\varphi) + \varepsilon^2 \int |\Delta \varphi|^2 + \text{something}.
%\]
%The first order term vanishes because $V$ solves the equations of motion.
%
%We look carefully at what the ``something'' is. We have
%\[
% (1 - (\Phi + \varepsilon \varphi, \Phi + \varepsilon \varphi))^2 = (1 - |\Phi|^2 - 2 \varepsilon (\varphi, \Phi) - \varepsilon ^2 |\varphi|^2)^2.
%\]
%The order $\varepsilon^2$ part is given by
%\[
% 4 \varepsilon^2 (\varphi, \Phi)^2 - 2 \varepsilon^2 (1 - |\phi|^2)|\varphi|^2
%\]
%So we find
%\[
% V(\Phi + \varepsilon \varphi) = V(\Phi) + \varepsilon^2 \int \left(|\nabla \varphi|^2 + \lambda (\varphi, \Phi)^2 - \frac{2\lambda}{4}(1 - |\Phi|^2)|\varphi|^2\right) \;\d x + O(\varepsilon^3).
%\]
%We now remember something about Green's functions, and the difference between massless and massive fields. We consider the ordinary Poisson equation
%\[
% - \Delta u = f.
%\]
%We can write down the solution to this in terms of Green's functions. In three dimensions, this is
%\[
% u(x) = \frac{1}{4\pi} \int \frac{f(y)}{|x - y|}\;\d y.
%\]
%This Green's function $\frac{1}{4\pi|x|}$ decays very slowly with $x$, and this is a sign that we have a massless field.
%
%If we change our integral equation instead to
%\[
% (-\Delta u + M^2 u) = f,
%\]
%then we find that
%\[
% u(x) + \frac{1}{4\pi} \int_{\R^3} \frac{e^{-M|x - y|}}{|x - y|} f(y) \;\d y.
%\]
%The Green's function is now exponentially decaying with $|x|$ if $M > 0$. This is characteristic for the behaviour of static Green's functions for massive fields.
%
%We now look at our change in $V$ above. We look at the quadratic fluctuation energy as $|x| \to \infty$. For $x$ very large, we have $(1 - |\Phi|^2) \to 0$ by the boundary conditions. Then we are left with the first two terms
%\[
% |\nabla \varphi|^2 + \lambda (\Phi, \varphi)^2.
%\]
%What does this mean? $\Phi$ and $\varphi$ are complex numbers, and in the second term, we are essentially projecting in the direction to $\Phi$. We write
%\[
% \varphi = \varphi^{||} + \varphi^{\perp}.
%\]
%Then we have
%\[
% \varphi^{||} \frac{(\Phi, \varphi)\Phi}{|\Phi|^2}.
%\]
%Thus, heuristically, this term is
%\[
% |\nabla \varphi^{||}|^2 + \lambda |\varphi^{||}|^2 + |\Delta \varphi^\perp|^2.
%\]
%So the parallel part behaves like a massive particle with quick decay, and the perpendicular part behaves like a massless particle with slow decay.

When we study vortices that are gauged, i.e.\ coupled to the electromagnetic field, we will see that all components are massive, and we don't have this problem. This is the Higgs mechanics.

This is all slightly heuristic, but is a good way to think about it.

\subsection{Abelian Higgs/Gauged Ginzburg--Landau vortices}
We now consider a theory where the complex scalar field $\Phi$ is coupled to a magnetic field. This is a $\U(1)$ gauge theory, with gauge potential given by a $1$-form
\[
  A = A_1 \;\d x^1 + A_2 \;\d x^2.
\]
The coupling between $\Phi$ and $A$ is given by \term{minimal coupling}. To do minimal coupling, we define the \term{covariant derivative}\index{$\D$}\index{$\D_A$}
\[
  \D \Phi = \D_A \Phi = \d \Phi - i A \Phi = \sum_{j = 1}^2 \D_j \Phi \;\d x^j
\]
To proceed, it is convenient to have a list of identities about the covariant derivative.
\begin{prop}
  If $f(x) \in \R$, then
  \[
    \D (f \Phi) = (\d f) \; \Phi + f\;\D \Phi.
  \]
  If $\Phi$ and $\Psi$ are complex scalar fields, then
  \[
    \d (\Phi, \Psi) = (\D \Phi, \Psi) + (\Phi, \D \Psi).
  \]
  In coordinates, these say
  \begin{align*}
    (\nabla_A)_j (f \Phi) &= (\partial_j f)\;\Phi + f\; \D_j \Phi\\
    \partial_j (\Phi, \Psi) &= (\D_j \Phi, \Psi) + (\Phi, \D_j \Psi).
  \end{align*}
\end{prop}
The proofs just involve writing all terms out. The second rule is analogous to fact that
\[
  \partial_k g(V, W) = g(\nabla_k V, W) + g(V, \nabla_k W)
\]
for the Levi-Civita connection.

The kinetic term is given by the magnetic field.
\begin{defi}[Magnetic field/curvature]\index{magnetic field}\index{curvature}
  The \emph{magnetic field}, or \emph{curvature} is given by
  \[
    B = \partial_1 A_2 - \partial_2 A_1.
  \]
  We can alternatively think of it as the 2-form
  \[
    \d A = B \;\d x^1 \wedge\d x^2.
  \]
\end{defi}
The formulation in terms of differential forms is convenient for computations, because we don't have to constrain ourselves to working in Cartesian coordinates --- often polar coordinates are more convenient.

\begin{prop}
  We have
  \[
    (\D_1 \D_2 - \D_2 \D_1) \Phi = - i B \Phi.
  \]
\end{prop}

The proof is again writing it out. Alternatively, we can express this without coordinates. We can extend $\D$ to act on $p$-forms by letting $A$ act as $A \wedge$. Then this result says
\begin{prop}
  \[
    \D \D \Phi = -i B \Phi.
  \]
\end{prop}

\begin{proof}
  \begin{align*}
    \D \D \Phi &= (\d - iA) (\d \Phi - iA\Phi)\\
    &= \d^2 \Phi - i \d (A \Phi) - iA\;\d \Phi + A \wedge A\;\Phi\\
    &= -i \d (A \Phi) - i A \;\d \Phi\\
    &= -i \d A\; \Phi + i A \;\d \Phi - i A\;\d \Phi\\
    &= -i (\d A)\; \Phi\\
    &= -i B \Phi
  \end{align*}
\end{proof}

The point of introducing the covariant derivative is that we can turn the global $\U(1)$ invariance into a local one. Previously, we had a global $\U(1)$ symmetry, where our field is unchanged when we replace $\Phi \mapsto \Phi e^{i \chi}$ for some constant $\chi \in \R$. With the covariant derivative, we can promote this to a \emph{global} symmetry.

Consider the simultaneous \term{gauge transformations}
\begin{align*}
  \Phi(x) &\mapsto e^{i\chi(x)} \Phi(x)\\
  A(x) &\mapsto A(x) + \d \chi(x).
\end{align*}
Then the covariant derivative of $\Phi$ transforms as
\begin{align*}
  (\d - iA) \Phi &\mapsto (\d - i (A + \d \chi)) (\Phi e^{i \chi}) \\
  &= (\d \Phi + i \Phi \d \chi - i (A + \d \chi) \Phi) e^{i\chi} \\
  &= e^{i \chi} (\d - iA) \Phi.
\end{align*}
Similarly, the magnetic field is also invariant under gauge transformations.

As a consequence, we can write down energy functionals that are invariant under these gauge transformations. In particular, we have
\[
  (\D \Phi, \D \Phi) \mapsto (e^{i\chi} \D \Phi, e^{i\chi} \D \Phi) = (\D \Phi, \D \Phi).
\]
So we can now write down the \term{gauged Ginzburg--Landau} \emph{energy}\index{Ginzburg--Landau!gauged}
\[
  V_\lambda(A, \Phi) = \frac{1}{2} \int_{\R^2} \left(B^2 + |\D \Phi|^2 + \frac{\lambda}{4} (1 - |\Phi|^2)^2\right)\;\d^2 x.
\]
This is then manifestly gauge invariant.
%where
%\[
% B^2 = \partial_1 A_2 - \partial_2 A_1,
%\]
%which can be shown to be gauge-invariant.
%
%Thus, $V(\lambda)$ is invariant under gauge transformations. In other words,
%\[
% V_\lambda(A + \d \chi, \Phi e^{i\chi}) = V_\lambda (A, \Phi).
%\]
%To proceed, it is convenient to have a list of identities about the covariant derivative.
%\begin{prop}
% If $f(x) \in \R$, then
% \[
% (\nabla_A)_j (f \Phi) = (\partial_j f) \Phi + f (\nabla_A)_j \Phi.
% \]
% If $\Phi$ and $\Psi$ are complex scalar fields, then
% \[
% \partial_j (\Phi, \Psi) = (\D_j \Phi, \Psi) + (\Phi, \D_j \Psi).
% \]
%\end{prop}
%The proofs just involve writing all terms out. The second rule is analogous to fact that
%\[
% \partial_k g(V, W) = g(\nabla_k V, W) + g(V, \nabla_k W)
%\]
%for the Levi-Civita connection.
%
%\begin{prop}
% We have
% \[
% -((\nabla_A)_1 (\nabla_A)_2 - (\nabla_A)_2 (\nabla_A)_1) \Phi = - i B \phi,
% \]
% where
% \[
% B = \partial_1 A_2 - \partial_2 A_1.
% \]
% In geometry, this \term{$B$} is known as the \term{curvature}, and in electromagnetism, this is the \term{magnetic field}.
%\end{prop}
%
%The proof is again writing it out.
%
%Our theory now has gauge invariance. In the ungauged Ginzburg--Landau theory, there was an invariance under a $\U(1)$ action $\Phi \mapsto \Phi e^{i\chi}$ for $\chi \in \R$ constant. This is \emph{global} gauge invariance, because it is constant everywhere.
%
%This invariance is now localized to
%\[
% \Phi(x) \mapsto \Phi(x) e^{i \chi(x)},
%\]
%where $\chi(x) \in \R$ is a smooth function, as long as our $A$ transforms accordingly as
%\[
% A \mapsto A + \d \chi.
%\]
%Then the covariant derivative changes as
%\[
% \D \Phi \mapsto (\D \Phi)e^{i \chi}.
%\]
%Finally, we check that the magnetic field is invariant,
%\[
% B \mapsto B.
%\]
%therefore the abelian Higgs energy functional
%\[
% V_\lambda (A, \Phi) = \frac{1}{2} \int \left(B^2 + |\D \Phi|^2 + \frac{\lambda}{4} (1 - |\Phi|^2)\right)\;\d^2 x
%\]
%satisfies
%\[
% V_\lambda(A + \d \chi, \Phi e^{i \chi}) = V_\lambda (A, \Phi)
%\]
%for all $\chi \in C^1(\R^2)$. So we now have an infinite degeneracy for all solutions.

As before, the equations of motion are given by the Euler--Lagrange equations. Varying $\Phi$, we obtain
\[
  - (\D_1^2 + \D_2^2) \Phi - \frac{\lambda}{2} (1 - |\Phi|^2)\Phi = 0.
\]
This is just like the previous vortex equation in the ungauged case, but since we have the covariant derivative, this is a coupled equation.

The other equations are given by
\begin{align*}
  \partial_2 B &= (i \Phi, \D_1 \Phi)\\
  -\partial_1 B &= (i \Phi, \D_2 \Phi).
\end{align*}
These are similar to Maxwell's equation, saying the curl of the magnetic filed is given by the current.

It is again an exercise to derive these. It is helpful to use the previous identities such as
\[
  \partial_j (\Phi, \Psi) = (\D_j \Psi, \Psi) + (\Phi, \D_j \Psi).
\]
So we get the integration by parts formula
\[
  \int_{\R^2} (\D_j \Phi, \Psi)\;\d^2 x = - \int (\Phi, \D_j \Psi)\;\d^2 x
\]
under suitable boundary conditions.

Recall that in the ungauged Gintzburg--Landau theory, we proved that we always had $|\Phi| \leq 1$. This is again true. The same proof works, with $\nabla$ replaced with $\D$. (It is literally the same proof. I just did copy-and-paste, find-and-replace)
\begin{lemma}
  At any maximum point of $|\Phi|$, we have $|\Phi| \leq 1$.
\end{lemma}

\begin{proof}
  Consider the function
  \[
    W(x) = 1 - |\Phi(x)|^2.
  \]
  Then we want to show that $W \geq 0$ when $W$ is minimized. We note that if $W$ is at a minima, then the Hessian matrix must have non-negative eigenvalues. So, taking the trace, we must have $\Delta W \geq 0$. Now we can compute $\Delta W$ directly. We have
  \begin{align*}
    \partial_j W &= -2 (\Phi, \D_j \Phi)\\
    \Delta W &= \partial_j^2 W \\
    &= - 2(\Phi, \D_j^2 \Phi) - 2(\D_j \Phi, \D_j \Phi)\\
    &= \lambda |\Phi|^2 W - 2 |\D \Phi|^2.
  \end{align*}
  Thus, we can rearrange to write this to say
  \[
    2 |\D \Phi|^2 + \Delta W = \lambda |\Phi|^2 W.
  \]
  But we know $2 |\D \Phi|^2 \geq 0$, and $\Delta W \geq 0$. So we must have $W \geq 0$.
\end{proof}

As before, this suggests we interpret $|\Phi|$ as an order parameter. This model was first used to describe superconductors. The matter can either be in a ``normal'' phase or a superconducting phase, and $|\Phi|$ measures how much we are in the superconducting phase.

Thus, in our model, far away from the vortices, we have $|\Phi| \approx 1$, and so we are mostly in the superconducting phase. The vortices represent a breakdown of the superconductivity. At the core of the vortices, we have $|\Phi| = 0$, and we are left with completely regular matter. Usually, this happens when we have a strong magnetic field. In general, magnetic field cannot penetrate the superconductor, but if it is strong enough, it will cause such breakdown in the superconductivity.

Similar to the ungauged case, for $\lambda > 0$, there exists basic vortex solutions of the form
\begin{align*}
  \Phi &= f_N(r) e^{iN\theta}\\
  A &= N \alpha_N(r) \;\d \theta.
\end{align*}
The boundary conditions $f_N, \alpha_N \to 1$ as $r \to \infty$ and $\to 0$ as $r \to 0$.

Let's say a few words about why these are sensible boundary conditions from the point of view of energy. We want
\[
  \lambda \int_{\R^2} (1 - |\Phi|)^2 < \infty,
\]
and this is possible only for $f_N \to 1$. What is less obvious is that we also need $\alpha_N \to 1$. We note that we have
\[
  \D_\theta \Phi = \frac{\partial \Phi}{\partial \theta} - i A_\theta \Phi = (iN f_N - iN \alpha_N f) e^{iN\theta}.
\]
We want this to $\to 0$ as $r \to \infty$. Since $f_N \to 1$, we also need $\alpha_N \to 1$.

The boundary conditions at $0$ can be argued as before, so that the functions are regular at $0$.

Let's calculate the topological charge. We have
\begin{align*}
  N &= \frac{1}{\pi} \int_{\R^2} j^0(\Phi)\;\d^2 x \\
  &= \lim_{R \to \infty} \frac{1}{2\pi} \oint_{|x| = R} (i \Phi, \d \Phi) \\
  &= \lim_{R \to \infty} \frac{1}{2\pi} \oint (if_N e^{iN\theta}, iN f_N e^{iN\theta})\;\d \theta\\
  &= \frac{1}{2\pi} \cdot N \lim_{R \to \infty} \int_0^{2\pi} f_N^2 \;\d \theta\\
  &= N
\end{align*}
Previously, we understood $N$ as the ``winding number'', and it measure how ``twisted'' our field was. However, we shall later see that there is an alternative interpretation of this $N$. Previously, in the sine-Gordon theory, we can alternatively think of $N$ as the number of kinks we've got. Similarly, here we can think of this vortex as a superposition of $N$ vortices at the origin. In the case of $\lambda = 1$, we will see that there are static solutions involving multiple vortices placed in different points in space.

We can compute the magnetic field as well. It is convenient to use the $\d A$ definition, as we are not working in Cartesian coordinates. We have
\[
  \d A = N \alpha_N'(r) \;\d r \wedge \d \theta = \frac{1}{r} N \alpha_N'\;\d x^1 \wedge \d x^2.
\]
Thus it follows that
\[
  B = \partial_1 A_2 - \partial_2 A_1 = \frac{N}{r} \alpha_N'.
\]
Now suppose that $|\Phi|^2 \to 1$ as $r \to \infty$, and also that $\D_\theta \Phi = o(r^{-1})$.
We now make the assumption that the angular part of the covariant derivative is $o(r^{-1})$ Then we find that
\[
  (\d \Phi)_\theta = i A_\theta \Phi + o(r^{-1}).
\]
Then if we integrate around a circular contour, since only the angular part contributes, we obtain
\[
  \oint_{|x| = R} (i \Phi, \d \Phi) = \oint_{|x| = R} (i \Phi, i A_\theta \Phi)\;\d \theta = \oint_{|x| = R} A + o(1).
\]
Note that here we are explicitly viewing $A$ as a differential form so that we can integrate it. We can then note that $|x| = R$ is the boundary of the disk $|x| \leq R$. So we can apply Stokes' theorem and obtain
\[
  \oint_{|x| = R} (i \Phi, \d \Phi) = \int_{|x| \leq R} \d A = \int_{|x| \leq R} B\;\d x^1 \;\d x^2.
\]

%
%Now suppose that the covariant derivative
%\[
% \D_\theta \Phi = (\nabla_A)_\theta \Phi = \left(\frac{\partial \Phi}{\partial \theta} - A_\theta \Phi\right) = (iN f_n - iN \alpha_N f_N) e^{iN\theta}
%\]
%is rapidly decreasing as $r \to \infty$, then we know
%\[
% A_\theta - \left(i \Phi, \frac{\partial \Phi}{\partial \theta}\right) \to 0.
%\]
%quickly as $r \to \infty$. Therefore we obtain
%\[
% \oint_{|x| = R} (i \Phi, \d \Phi) = \oint_{|x| = R} \left(i \Phi, \frac{\partial \Phi}{\partial \theta}\right)\;\d \theta = \oint_{|x| = R} A_\theta \;\d \theta + o(1).
%\]
%We now apply Stokes theorem. We note that $A$ only has a $\theta$ direction. So we have
%\[
% A_\theta \;\d \theta = A_1 \;\d x^1 + A_2\;\d x^2.
%\]
%Thus we obtain
%\[
% \oint_{|x| = R} A_1 \;\d x^1 + A_2\;\d x^2 = \int_{|x| \leq R} (\partial_1 A_2 - \partial_2 A_1)\;\d x^1 \wedge \d x^2 = \int_{|x| \leq R} B\;\d x^1 \;\d x^2.
%\]
Now we let $R \to \infty$. Then we get
\[
  \int_{\R^2} B\;\d x^1 \;\d x^2 = \lim_{R \to \infty} \oint_{|x| = R} A = \lim_{R \to \infty} \oint (i \Phi, \d \Phi) = 2 \int_{\R^2} j^0.
\]
%If we analyze our derivation of this, it didn't exactly require that we had a radial vortex. It only required that
%\[
% |\D \Phi| \leq \frac{\text{constant}}{r^{1 + \varepsilon}}
%\]
%for some $\varepsilon > 0$.

Physically, what this tells us then is that there is a relation between the topological winding number and the magnetic flux. This is a common property of topological gauge theories. In mathematics, this is already well known --- it is the fact that we can compute characteristic classes of vector bundles by integrating the curvature, as discovered by Chern.

\subsubsection*{Behaviour as $r \to 0$}
We saw last time that for reasons of regularity, it was necessary that $f_N, \alpha_N \to 0$ as $r \to 0$.

But actually, we must have $f_N \sim r^N$ as $r \to 0$. This has a consequence that $\Phi \sim (r e^{i\theta})^N = z^N$. So the local appearance of the vortex is the zero of an analytic function with multiplicity $N$.

To see this, we need to compute that the Euler--Lagrange equations are
\[
  -f_N'(R) - \frac{1}{r} f_N'(r) + \frac{(N - \alpha_N)^2}{r^2} f_N = \frac{\lambda}{2} f_N(1 - f_N^2).
\]
With the boundary condition that $f_N$ and $\alpha_N$ vanish at $r = 0$, we can approximate this locally as
\[
  - f_N'' - \frac{1}{r} f_N' + \frac{N^2}{r^2} f_N = 0.
\]
since we have a $\frac{1}{r^2}$ on the left hand side. This equation is homogeneous, and then the solutions are just
\[
  f_N = r^{\pm N}.
\]
So for regularity, we want the one that $\to 0$. So we take $f_N \sim r^N$.

\subsection{Bogomolny/self-dual vortices and Taubes' theorem}
As mentioned, we can think of our previous solution as a collection of $N$ vortices all superposed at the origin. Is it possible to have separated vortices all over the plane? Naively, we would expect that the vortices exert forces on each other, and so we don't get a static solution. However, it turns out in the $\lambda = 1$ case, we do get static, separated solutions.

This is not obvious, and the proof requires some serious analysis. We will not do the analysis, because we don't care about all that Sobolev nonsense. However, we will do all the non-hard-analysis part. In particular, we will obtain Bogomolny bounds as we did in the sine-Gordon case, and reduce the problem to finding solutions to a single scalar PDE, which we can let analysts tackle with their tools.

Recall that when we did the sine-Gordon kinks, we wanted
\[
  \theta'' = \sin \theta,
\]
with boundary conditions $\theta(x) \to 0$ or $2\pi$ as $x \to \pm \infty$. The only solutions we found were
\[
  \theta_K (x - X)
\]
for any $X \in \R$. This $X$ is interpreted as the location of the kink. So the moduli space of solutions is $\mathcal{M} = \R$.

We shall get a similar but more interesting description for the $\lambda = 1$ vortices. This time, the moduli space will be $\C^N$, given by $N$ complex parameters describing the equilibrium.

\begin{thm}[Taubes' theorem]
  For $\lambda = 1$, the space of (gauge equivalence classes of) solutions of the Euler--Lagrangian equations $\delta V_1 = 0$ with winding number $N$ is $\mathcal{M} \cong \C^N$.
  
  To be precise, given $N \in \N$ and an unordered set of points $\{Z_1, \cdots, Z_N\}$, there exists a solution $A(x; Z_1, \cdots, Z_N)$ and $\Phi(x; Z_1, \cdots, Z_n)$ which solves the Euler-Lagrange equations $\delta V = 0$, and also
  \[
    B = \frac{1}{2} (1 - |\Phi|^2),\quad \D_1 \Phi + i \D_2 \Phi = 0.
  \]
  These are known as the \term{Bogomolny equations}. Moreover, $\Phi$ has exactly $N$ zeroes $Z_1, \cdots, Z_n$ counted with multiplicity, where
  \[
    \Phi(x; Z_1, \cdots, Z_N) \sim c_j (z - Z_j)^{n_j}
  \]
  as $z \to Z_j$, where $n_j = |\{k: Z_k = Z_j\}|$ is the multiplicity.

  This solution is unique up to gauge equivalence invariance.

  Furthermore,
  \[
    V_1(A(\ph, Z_1, \cdots, Z_N), \Phi(\ph; Z_1, \cdots, Z_n)) = \pi N\tag{$*$}
  \]
  and
  \[
    \frac{1}{2\pi} \int_{\R^2} B\;\d x^1 \;\d x^2 = N = \text{winding number}.
  \]
\end{thm}
Note that it is not immediately clear from our description that the moduli space is $\C^N$. It looks more like $\C^N$ quotiented out by the action of $S^N$. However, we will later see that the resulting quotient is still isomorphic to $\C^N$.

There is a lot to be said about this theorem. The equation $(*)$ tells us the energy is just a function of the number of particles, and does not depend on where they are. This means there is no force between the vortices. In situations like this, people say the Bogomolny bound is saturated. The final statement tells us topology is what is driving the vortices, as we have already seen.

Note that the Euler--Lagrange equations themselves are second-order equations. However, the Bogomolny equations are \emph{first order}. In general, this is a mathematical signature that suggests these interesting structures are present.

We'll discuss three crucial ingredients in this, but we will not complete the proof, because it is too much analysis. The proof can be found in a book by \emph{Jaffe and Taubes}

\subsubsection*{Holomorphic structure}
When we get these structures, there is often some complex analysis lurking behind. We can explicitly write the second Bogomolny equation as
\[
  \D_1 \Phi + i \D_2 \Phi = \frac{\partial \Phi}{\partial x_1} + \frac{\partial \Phi}{\partial x_2} - i(A_1 + i A_2) \Phi = 0.
\]
Recall that in complex analysis, holomorphic functions are characterized by
\[
  \frac{\partial f}{\partial \bar{z}} = \frac{1}{2} \left(\frac{\partial f}{\partial x^1} + i \frac{\partial f}{\partial x^2}\right) = 0.
\]
So we think of the second Bogomolny equation as a covariant Cauchy--Riemann equations. It is possible to convert this into the Cauchy--Riemann equations to deduce the local behaviour at $\Phi$.

To do so, we write
\[
  \Phi = e^{\omega f}.
\]
Then
\begin{align*}
  \frac{\partial f}{\partial \bar{z}} &= e^{-\omega} \left(\frac{\partial \Phi}{\partial \bar{z}} - \frac{\partial \omega}{\partial \bar{z}} \Phi\right)\\
  &= e^{-\omega} \left(\frac{i(A_i + i A_2)}{2} - \frac{\partial \omega}{\partial \bar{z}}\right)\Phi.
\end{align*}
This is equal to $0$ if $\omega$ satisfies
\[
  \frac{\partial \omega}{\partial \bar{z}} = i \frac{A_1 + iA_2}{2}.
\]
So the question is --- can we solve this equation? It turns out we can always solve this equation, and there is an explicit formula for this. In general, for the equation
\[
  \frac{\partial w}{\partial \bar{z}} = \beta,
\]
the solution is given by
\[
  \omega(z) = \frac{1}{2\pi i} \int_{|w| < r} \frac{\beta(w)}{w - z}\;\d w \wedge \d \bar{w}.
\]
A proof can be found in a book on algebraic geometry, namely \emph{Griffiths and Harris}. Fortunately, the proof of this result is on page 5, and you do not have to go through hundreds of pages of algebraic geometry to get to this result!

Then we can write
\[
  \Phi = e^{\omega} f
\]
where $f$ is holomorphic. Since $e^{\omega}$ is never zero, we can apply all our knowledge of holomorphic functions to $f$, and deduce that $\Phi$ has isolated zeroes, where $\Phi \sim (z - Z_j)^{n_j}$ for some power $n_j$.

\subsubsection*{The Bogomolny equations}
We'll now show that $(A, \Phi)$ satisfies $V_1(A, \Phi) = \pi N$ iff it satisfies the Bogomolny equation, i.e.
\[
  \D_1 \Phi + i \D_2 \Phi = 0,\quad B = \frac{1}{2} (1 - |\Phi|^2).
\]
To see this, recall for the sine-Gordon kink, we wrote the energy as
\begin{align*}
  E &= \int_{-\infty}^\infty \left(\frac{1}{2} \theta_x^2 + (1 - \cos \theta)\right) \;\d x \\
  &= \frac{1}{2} \int_{-\infty}^\infty \left(\theta_x^2 + 4 \sin^2 \frac{\theta}{2}\right) \;\d x\\
  &= \frac{1}{2} \int_{-\infty}^\infty \left(\left(\theta_x - 2 \sin \frac{\theta}{2}\right)^2 + 4 \theta_x \sin \frac{\theta}{2}\right)\;\d x\\
  &= \frac{1}{2} \int_{-\infty}^\infty \left(\theta_x - 2 \sin \frac{\theta}{2}\right)^2\;\d \theta + \int \frac{\partial}{\partial x}\left(-8 \cos \frac{\theta}{2}\right)\;\d x\\
  &= \frac{1}{2}\int_{-\infty}^\infty \left(\theta_x - 2 \sin \frac{\theta}{2}\right)^2\;\d \theta + \left(-4\cos \frac{\theta(+\infty)}{2} + 4 \cos \frac{\theta(-\infty)}{2}\right).
\end{align*}
We then use the kink asymptotic boundary conditions to obtain, say, $\theta(+\infty) = 2\pi$ and $\theta(-\infty) = 0$. So the boundary terms gives $8$. Thus, subject to these boundary conditions, we can write the sine-Gordon energy as
\[
  E = \frac{1}{2}\int \left(\theta_x - 2 \sin \frac{\theta}{2}\right)^2 + 8.
\]
Thus, if we try to minimize the energy, then we know the minimum would be at least $8$, and if we could solve the first-order equation $\theta_x = 2 \sin \frac{\theta}{2}$, then the minimum would be exactly $8$. This is the solution we found, and the solution is exactly the kink solution we found. Moreover, the solutions are all of the form
\[
  \theta(x) = \theta_K(x - X),\quad \theta_k(x) = 4 \arctan e^x.
\]
Thus, we have shown that the minimum energy is $8$, and the minimizers are all of this form, parameterized by $X \in \R$.

We want to the do something similar for the Ginzburg--Landau theory
\begin{lemma}
  We have
  \[
    V_1(A, \Phi) = \frac{1}{2} \int_{\R^2} \left(\left(B - \frac{1}{2}(1 - |\Phi|^2)\right)^2 + 4 |\bar{\partial}_A \Phi|^2\right)\;\d x^1\;\d x^2 + \pi N,
  \]
  where
  \[
    \bar{\partial}_A \Phi = \frac{1}{2} (\D_1 \Phi + i \D_2 \Phi).
  \]
\end{lemma}
It is clear that the desired result follows from this.

\begin{proof}
  We complete the square and obtain
  \[
    V_1(A, \Phi) = \frac{1}{2} \int \left(\left(B - \frac{1}{2}(1 - |\Phi|^2)\right)^2 + B(1 - |\Phi|^2) + |\D_1\Phi|^2 + |\D_2 \Phi|^2\right)\;\d^2 x
  \]
  We now dissect the terms one by one. We first use the definition of $B = \d A$ and integration by parts to obtain
  \[
    \int_{\R^2} (1 - |\Phi|^2) \;\d A = - \int_{\R^2} \d (1 - |\Phi|^2) \wedge A = 2 \int_{\R^2} (\Phi, \D \Phi) \wedge A.
  \]
  Alternatively, we can explicitly write
  \begin{align*}
    \int_{\R^2} B(1 - |\Phi|^2) \;\d^2 x &= \int_{\R^2} (\partial_1 A_2 - \partial_2 A_1) (1 - |\Phi|^2)\;\d ^2x\\
    &= \int_{\R^2} (A_2 \partial_1 |\Phi|^2 - A_1 \partial_2 |\Phi|^2) \;\d^2 x \\
    &= 2\int_{\R^2} A_2 (\Phi, \D_1 \Phi) - A_1 (\Phi, \D_2 \Phi).
  \end{align*}
  Ultimately, we want to obtain something that looks like $|\bar{\partial}_A \Phi|^2$. We can write this out as
  \[
    (\D_1 \Phi + i \D_2 \Phi, \D_1 \Phi + i \D_2 \Phi) = |\D_1 \Phi|^2 + |\D_2 \Phi|^2 + 2(\D_1 \Phi, i \D_2 \Phi).
  \]
  We note that $i\Phi$ and $\Phi$ are always orthogonal, and $A_i$ is always a real coefficient. So we can write
  \begin{align*}
    (\D_1 \Phi, i \D_2 \Phi) &= (\partial_1 \Phi - i A_1 \Phi, i \partial_2 \Phi + A_2 \Phi) \\
    &= (\partial_1 \Phi, i \partial _2 \Phi) + A_2 (\Phi, \partial_1 \Phi) - A_1 (\Phi, \partial_2 \Phi)
  \end{align*}
  We now use again the fact that $(\Phi, i\Phi) = 0$ to replace the usual derivatives with the covariant derivatives. So we have
  \[
    (\D_1 \Phi, i \D_2 \Phi) = (\partial_1 \Phi, i \partial_2 \Phi) + A_2 (\Phi, \D_1 \Phi) - A_1 (\Phi, \D_2 \Phi).
  \]
  This tells us we have
  \[
    \int \left(B(1 - |\Phi|^2) + |\D_1 \Phi|^2 + |\D_2 \Phi|^2\right)\;\d^2 x = \int \left(4 |\bar{\partial}_A \Phi|^2 + 2 (\partial_1 \Phi, i \partial_2 \Phi)\right)\;\d^2 x.
  \]
  It then remains to show that $(\partial_1 \Phi, i \partial_2 \Phi) = j^0(\Phi)$. But we just write
% This is just like what we had above! So we get
% \begin{align*}
% V_1(A, \Phi) &= \frac{1}{2} \int \left(\left(B - \frac{1}{2}(1 - |\Phi|^2)\right)^2 + |\D_1 \Phi|^2 + |\D_2 \Phi|^2 + 2 (\D_1 \Phi, i \D_2 \Phi) - 2 (\partial_1 \Phi, i \partial_2 \Phi)\right) \;\d^2 x\\
% &= \frac{1}{2} \int \left(\left(B - \frac{1}{2} (1 - |\Phi|^2)\right)^2 + 4 |\bar{\partial}_A \Phi|^2 + 2 j^0(\Phi)\right)\;\d^2 x,
% \end{align*}
% since
  \begin{align*}
    (\partial_1 \Phi_1 + i \partial_1 \Phi_2, - \partial_2 \Phi_2 + i \partial_2 \Phi_1) &= - (\partial_1 \Phi_1, \partial_2 \Phi_2) + (\partial_1 \Phi_2, \partial_2 \Phi_1)\\
    &= -j^0(\Phi)\\
    &= - \det
    \begin{pmatrix}
      \partial_1 \Phi_1 & \partial_2 \Phi_1\\
      \partial_1 \Phi_2 & \partial_2 \Phi_2
    \end{pmatrix}
  \end{align*}
  Then we are done.
\end{proof}

\begin{cor}
  For any $(A, \Phi)$ with winding number $N$, we always have $V_1(A, \Phi) \geq \pi N$, and those $(A, \Phi)$ that achieve this bound are exactly those that satisfy
  \[
    \bar{\partial}_A \Phi = 0,\quad B = \frac{1}{2} (1 - |\Phi|^2).
  \]
\end{cor}

\subsubsection*{Reduction to scalar equation}
The remaining part of Taubes' theorem is to prove the existence of solutions to these equations, and that they are classified by $N$ unordered complex numbers. This is the main analytic content of the theorem.

To do so, we reduce the two Bogomolny equations into a scalar equation. Note that we have
\[
  \D_1 \Phi + i \D_2 \Phi = (\partial_1 \Phi + i \partial_2 \Phi) - i (A_1 + i A_2) \Phi = 0.
\]
So we can write
\[
  A_1 + i A_2 = - i (\partial_1 + i \partial_2) \log \Phi.
\]
Thus, once we've got $\Phi$, we can get $A_1$ and $A_2$.

The next step is to use gauge invariance. Under gauge invariance, we can fix the phase of $\Phi$ to anything we want. We write
\[
  \Phi = e^{\frac{1}{2} (u + i \theta)}.
\]
Then $|\Phi|^2 = e^u$.

We might think we can get rid of $\theta$ completely. However, this is not quite true, since the argument is not well-defined at a zero of $\Phi$, and in general we cannot get rid of $\theta$ by a smooth gauge transformation. But since
\[
  \Phi \sim c_j (z - Z_j)^{n_j}
\]
near $Z_j$, we can expect we can make $\theta$ look like
\[
  \theta = 2 \sum_{j = 1}^N \arg (z - Z_j).
\]
We will assume we can indeed do so. Then we have
\[
  A_1 = \frac{1}{2}(\partial_2 u + \partial_1 \theta),\quad A_2 = - \frac{1}{2} (\partial_1 u - \partial_2 \theta).
\]
We have now solved for $A$ using the first Bogomolny equation. We then use this to work out $B$ and obtain a scalar equation for $u$ by the second Bogomolny equation.

\begin{thm}
  The Bogomolny equation $B = \frac{1}{2} (1 - |\Phi|^2)$ is equivalent to the scalar equation for u
  \[
    -\Delta u + (e^u - 1) = -4\pi \sum_{j = 1}^N \delta_{Z_j}.
  \]
  This is known as \term{Taubes' equation}.
\end{thm}

\begin{proof}
  We have
  \begin{align*}
    B &= \partial_1 A_2 - \partial_2 A_1 \\
    &= -\frac{1}{2} \partial_1^2 u - \frac{1}{2} \partial_2^2 u + \frac{1}{2} (\partial_1 \partial_2 - \partial_2 \partial_1) \theta\\
    &= - \frac{1}{2} \Delta u + \frac{1}{2} (\partial_1 \partial_2 - \partial_2 \partial_1) \theta.
  \end{align*}
  We might think the second term vanishes identically, but that is not true. Our $\theta$ has some singularities, and so that expression is not going to vanish at the singularities. The precise statement is that $(\partial_1 \partial_2 - \partial_2 \partial_1) \theta$ is a distribution supported at the points $Z_j$.

  To figure out what it is, we have to integrate
  \[
    \int_{|z - Z_j| \leq \varepsilon} (\partial_1 \partial_2 - \partial_2 \partial_1) \theta = \int_{|z - Z_j| = \varepsilon} \partial_1 \theta \;\d x^1 + \partial_2 \theta \;\d x^2 = \oint_{|z - Z_j| = \varepsilon} \;\d \theta = 4\pi n_j,
  \]
  where $n_j$ is the multiplicity of the zero. Thus, we deduce that
  \[
    (\partial_1 \partial_2 - \partial_2 \partial_1) \theta = 2\pi \sum \delta_{Z_j}.
  \]
  But then we are done!
\end{proof}
We can think of this $u$ as a non-linear combination of fundamental solutions to the Laplacian. Near the $\delta$ functions, the $e^u - 1$ term doesn't contribute much, and the solution looks like the familiar fundamental solutions to the Laplacian. However, far away from the singularities, $e^u - 1$ forces $u$ to tend to $0$ at infinity, instead of growing to infinity.

Taubes proved this has a unique solution, which is smooth on $\R^2 \setminus \{Z_j\}$, with logarithmic singularities at $Z_j$, and such that $u \to 0$ as $|z| \to \infty$. Also, $u < 0$.

It is an exercise to check that the Bogomolny equations imply the second-order Euler--Lagrange equation.

For example, differentiating the first Bogomolny equation gives
\[
  \partial_1 B = - (\Phi, \D_1 \Phi) = (\Phi, i \D_2 \Phi).
\]
We can similarly do this for the for the sine-Gordon theory.

\subsection{Physics of vortices}
Recall we began with the ungauged Ginzberg--Landau theory, and realized the solitons didn't have finite energy. We then added a gauge field, and the problem went away by itself. When we did this, we heuristically justified it by saying it ``gave mass'' to the angular component. However, there is another way of understanding it. And this doesn't require \emph{any} detailed information about the theory at all.

Suppose we work in $d$ space dimensions. Then a general scalar field $\Phi: \R^d \to \R^\ell$ has energy functional given by
\[
  \int_{\R^d} \left(\frac{1}{2} |\Delta \Phi|^2 + U(\Phi)\right)\;\d^d x.
\]
for some $U$.

\begin{thm}[Derrick's scaling argument]\index{Derrick's scaling argument}
  Consider  a field theory in $d$-dimensions with energy functional
  \[
    E[\Phi] = \int_{\R^d}\left(\vphantom{\frac{1}{2}}\right.\underbrace{\frac{1}{2} |\nabla \Phi|^2}_{T} + \underbrace{U(\Phi)\vphantom{\frac{1}{2}}}_W\left.\vphantom{\frac{1}{2}}\right)\;\d^d x.
  \]
  \begin{itemize}
    \item If $d = 1$, then any stationary point must satisfy
      \[
        T = W.
      \]
    \item If $d = 2$, then all stationary points satisfy $W = 0$.
    \item If $d \geq 3$, then all stationary points have $T = W = 0$, i.e.\ $\Phi$ is constant.
  \end{itemize}
\end{thm}
This forbids the existence of solitons in high dimensions.

\begin{proof}
  Suppose $\Phi$ were such a stationary point. Then for any variation $\Phi_\lambda$ of $\Phi$ such that $\Phi = \Phi_1$, we have
  \[
    \left.\frac{\d}{\d \lambda}\right|_{\lambda = 1}E[\Phi_\lambda] = 0.
  \]
  Consider the particular variation given by
  \[
    \Phi_\lambda(\mathbf{x}) = \Phi(\lambda \mathbf{x}).
  \]
  Then we have
  \[
    W[\Phi_\lambda] = \int_{\R^d} U(\Phi_\lambda(\mathbf{x})) \;\d^d x = \lambda^{-d} \int_{\R^d} U(\Phi(\lambda \mathbf{x}))\;\d^d (\lambda x) = \lambda^{-d} W[\Phi].
  \]
  On the other hand, since $T$ contains two derivatives, scaling the derivatives as well gives us
  \[
    W[\Phi_\lambda] = \lambda^{2 -d} W[\Phi].
  \]
  Thus, we find
  \[
    E[\Phi_\lambda] = \lambda^{2 - d}T[\Phi] + \lambda^{-d} W[\Phi].
  \]
  Differentiating and setting $\lambda = 1$, we see that we need
  \[
    (2 - d) T[\Phi] - d W[\Phi] = 0.
  \]
  Then the results in different dimensions follow.
\end{proof}
%We can look at vortices for general $\lambda > 0$. We shall begin by some high-level argument involving \term{Derrick's scaling argument}. For scalar fields $\phi: \R^d \to \R^\ell$, suppose we have an energy functional
%\[
%  \int_{\R^d} \left(\frac{1}{2} |\Delta \phi|^2 + U(\phi)\right)\;\d^d x.
%\]
%We assume that the potential $U \geq 0$, and that there exists a finite-energy smooth equilibrium $\phi$. In other words, if $\phi_\lambda$ is a smooth family of field configurations with $\phi_\lambda |_{\lambda = 1} = \phi$, then
%\[
%  \left.\frac{\d}{\d \lambda} V(\phi_\lambda) \right|_{\lambda = 1} = 0.
%\]
%We now want to ask if these things exist.
%
%We choose our variation by
%\[
%  \phi_\lambda(x) = \phi(\lambda x).
%\]
%We can evaluate
%\begin{align*}
%  V(\phi_\lambda) &= \int_{\R^d} \left(\frac{1}{2} \lambda^2|(\Delta \phi)(\lambda x)|^2 + U(\phi_\lambda)\right)\;\d^d x\\
%  &= \int_{\R^d} \left(\frac{1}{2} \lambda^2 |\Delta \phi(y)|^2  + U(\phi(y))\right)\;\lambda^{-d} \;\d^d y,
%\end{align*}
%where we substituted in $y = \lambda x$. We write
%\[
%  T = \int_{\R^d} \frac{1}{2} |\Delta \phi|^2 \;\d^d x,\quad W = \int_{\R^d} U(\phi)\;\d^d x.
%\]
%Then we find that
%\[
%  V(\phi_\lambda)  = \lambda^{2 - d} T + \lambda^{-d} W.
%\]
%Here $T$ and $W$ are constants completely determined by the soliton solution, and is in particular independent of $\lambda$. So we find
%\[
%  \left.\frac{\d}{\d \lambda} (V(\phi_\lambda))\right|_{\lambda = 1} = (2 - d) T - d W = 0.
%\]
%Now if we want interesting, non-vacuum solutions, then $T$ and $W$ are non-zero.
%
%If $d = 1$, then there is no problem. We just get that we must have
%\[
%  T = W.
%\]
%However, if $d \geq 3$, then we immediately get an impossibility, as both coefficients are negative. So this happens only if $T = W = 0$. So $\phi$ is constant.
%
%If $d = 2$, then this is a bit border-line. Then the kinetic energy term doesn't matter. Then we get a solution only if $W = 0$. There are some interesting solitons satisfying this, if the vacuum states are degenerate enough.

The $d = 2$ is rather interesting. We can still get interesting soliton theories if we have sufficiently degenerate vacuum states.
\begin{eg}
  In $d = 2$, we can take $\ell = 3$ and
  \[
    W(\phi) = (1 - |\phi|^2)^2.
  \]
  Then the set $W = 0$ is given by the unit sphere $S^2 \subseteq \R^3$. This is a \term{$\sigma$-model}, and there is a large class of such maps generated by rational functions.
\end{eg}

Derrick's scaling argument is not only a mathematical trick. We can also interpret it physically. Increasing the $\lambda$ corresponds to ``collapsing'' down the field. Then we see that in $d \geq 3$, both the kinetic and potential terms favour collapsing of the field. However, in $d = 1$, the kinetic term wants the field to expand, and the potential term wants the field to collapse. If these two energies agree, then these forces perfectly balance, and stationary solitons exist.

Now let's consider the case of a gauge theory. To understand this, we need to know how we want the gauge field to transform. There as several ways of figuring this out.

For example, we can insist the covariant derivative $\D_j \Phi_\lambda$ to scale as a whole. Since
\[
  \partial_j \Phi_\lambda = \lambda (\partial_j \Phi)(\lambda x),
\]
we would want $A_j$ to scale as
\[
  (A_j)_\lambda(x) = \lambda A_j(\lambda x).
\]
We can also do this more geometrically. Consider the function
\begin{align*}
  \chi_\lambda: \R^d &\to \R^d\\
  x &\mapsto \lambda x.
\end{align*}
Then our previous transformations were just given by pulling back along $\chi_\lambda$. Since $A$ is a $1$-form, it pulls back as
\[
  \chi_\lambda^* (A_j \;\d x^j) = \lambda A_j (\lambda x) \;\d x^j.
\]
Then since $B = \d A$, the curvature must scale as
\[
  B_\lambda(x) = \lambda^2 B(\lambda x).
\]
Thus, we can obtain a gauged version of Derrick's scaling argument.

Since we don't want to develop gauge theory in higher dimensions, we will restrict our attention to the Ginzburg--Landau model. Since we already used the letter $\lambda$, we will denote the scaling parameter by $\mu$. We have
\begin{align*}
  V_\lambda(A_\mu, \Phi_\mu) &= \frac{1}{2} \int \left(\mu^4 B(\mu x)^2 + \mu^2 |\D \Phi(\mu x)|^2 + \frac{\lambda}{4} (1 - |\Phi(\mu x)|^2)^2\right) \frac{1}{\mu^2}\;\d^2 (\mu x)\\
  &= \frac{1}{2} \int \left(\mu^2 B^2(y) + |\D \Phi(y)|^2 + \frac{\lambda}{4 \mu^2} (1 - |\Phi(y)|^2)^2\right)\;\d^2 y.
\end{align*}
Again, the kinetic term is scale invariant, but the gauge term counteracts the potential term. We can find the derivative to be
\[
  \left.\frac{\d}{\d \mu}\right|_{\mu = 1} V_\lambda(A_\mu, \Phi_\mu) = \int \left(B^2 - \frac{\lambda}{4}(1 - |\Phi|^2)^2\right)\;\d^2 y.
\]
Thus, for a soliton, we must have
\[
  \int B^2\;\d^2 x = \frac{\lambda}{4} \int_{\R^2} (1 - |\Phi|^2)^2 \;\d^2 x.
\]
Such solutions exist, and we see that this is because they are stabilized by the magnetic field. Note that in the case $\lambda = 1$, this equation is just the integral form of one of the Bogomolny equation!

We will eventually want to work with theories in higher dimensions, and Derrick's scaling argument shows that this isn't going to be very successful, even if we introduce gauge fields. There are different ways to get around Derrick's argument. In the Skyrme model, which is what we are going to study in the next chapter, we will do so by introducing some higher derivative terms. In particular, by introducing fourth derivatives, we will have a term that scales as $\lambda^{4 - d}$, and this allows us to stay safe in three dimensions, 

\subsubsection*{Scattering of vortices}
Derrick's scaling argument suggests that vortices can exist if $\lambda > 0$. However, as we previously discussed, for $\lambda \not= 1$, there are forces between vortices in general, and we don't get static, separated vortices. By doing numerical simulations, we find that when $\lambda < 1$, the vortices attract. When $\lambda > 1$, the vortices repel. Thus, when $\lambda > 1$, the symmetric vortices with $N > 1$ are unstable, as they want to break up into multiple single vortices.

We can talk a bit more about the $\lambda = 1$ case, where we have static multi-vortices. For example, in $N = 2$, the solutions are parametrized by pairs of points in $\C$, up to
\[
  (Z_1, Z_2) \sim (Z_2, Z_1).
\]
We said the moduli space is $\C^2$, and this is indeed true. However, $Z_1$ and $Z_2$ are no longer good coordinates for this space. Instead, good coordinates on the space $\mathcal{M} = \mathcal{M}_2$ is given by some functions symmetric in $Z_1$ and $Z_2$. One valid choice is
\[
  Q = Z_1 + Z_2,\quad P = Z_1 Z_2.
\]
In general, for higher dimensions, we should use the \term{elementary symmetric polynomials} in $Z_1, \cdots, Z_j$ as our coordinates. 

Now suppose we set our vortices in motion. For convenience, we fix the center of mass so that $Q(t) = 0$. We can then write $P$ as
\[
  P = - Z_1^2.
\]
If we do some simulations, we find that sometimes, after they collide, they scatter off in $90^\circ$. This is the \term{$90^\circ$ scattering phenomenon}, and holds for other $\lambda$ as well.

In terms of our coordinates, this corresponds to $Z_1 \mapsto i Z_1$, $Z_2 = - Z_1$. Note that at the point when they collide, we lose track of which vortex is which, which is 

Similar to the $\phi^4$ kinks, we can obtain effective Lagrangians for the effective dynamics of these vortices. However, this is much more complicated.

\subsection{Jackiw--Pi vortices}
So far, we have been thinking about electromagnetism, using the abelian Higgs model. There is a different system that is useful in condensed matter physics. We look at vortices in Chern--Simons--Higgs theory. This is a different Lagrangian with no relativity any more. Thus, instead of having the curvature term, we have the Chern--Simons Lagrangian. We again work in two-dimensions, with the Lagrangian density given by
\[
  \mathcal{L} = \frac{\kappa}{4} \varepsilon^{\mu\nu\lambda} A_\mu F_{\nu\lambda} - (i \Phi, \D_0 \Phi) - \frac{1}{2} | \D \Phi|^2 + \frac{1}{2\kappa} |\Phi|^4,
\]
where we again have
\[
  \D_0 \Phi = \frac{\partial \Phi}{\partial t} - i A_0 \Phi.
\]
The first term is the \term{Chern--Simons term}, while the remaining is the Schr\"odinger action with a $|\Phi|^4$ potential term.

Varying with respect to $\Phi$, the Euler--Lagrange equations gives the Schr\"odinger equation
\[
  i D_0 \Phi + \frac{1}{2} \D_j^2 \Phi + \frac{1}{\kappa} |\Phi|^2 \Phi = 0.
\]
If we take the variation with respect to $A_0$ instead, then we have
\[
  \kappa B + |\Phi|^2 = 0.
\]
This is a bit funny, because it looks more like a constraint than an evolution equation. This is characteristic of Chern--Simons theories.

The other equations give conditions like
\begin{align*}
  \partial_1 A_0 &= \partial_0 A_1 + \frac{1}{\kappa} (i \Phi, \D_2 \Phi)\\
  \partial_2 A_0 &= \partial_0 A_2 - \frac{1}{\kappa} (i \Phi, \D_1 \Phi).
\end{align*}
This is a bit odd, compared to Maxwell theory, because this is directly relating the electromagnetic field to the current, rather than its derivative.

For static solutions, we need
\[
  \partial_1 A_0 = \frac{1}{\kappa} (i \Phi, \D_2 \Phi).
\]
To solve this, we assume that the field again satisfies
\[
  \D_1 \Phi + i \D_2 \Phi = 0.
\]
Then we can write
\[
  \partial_1 A_0 (i \Phi, \D_2 \Phi) = - \frac{1}{\kappa} \partial_1 \frac{|\Phi|^2}{2}.
\]
We can then integrate this to obtain
\[
  A_0 = -\frac{|\Phi|62}{ 2 \kappa}.
\]
This will solve the last two equations, subject to the holomorphicity assumption.

We can then look at the other two equations, and see how we can solve those. We put this into the Schr\"odinger equation to obtain
\[
  A_0 \Phi + \frac{1}{2} \D_j^2 \Phi + \frac{1}{\kappa} |\Phi|^2 \Phi = 0.
\]
Substituting in $A_0$, we obtain
\[
  \left(- \D_j^2 \Phi = \frac{|\Phi|^2}{\kappa} \right)\Phi = 0.
\]
Let's then see if this makes sense. We need to see whether this can be consistent with the holomorphicity condition. The answer is that it can be, if we have the equation $\kappa B + |\Phi|^2 = 0$. We can calculate
\[
  \D_j^2 \Phi = \D_1^2 \Phi + \D_2^2 \Phi = -i (\D_1 \D_2 - \D_2 \D_1) \Phi = - B \Phi = \frac{1}{\kappa} |\Phi|^2 \Phi.
\]
Then this equation is exactly what we wanted.

So the conclusion is that we can generate vortex solutions by solving
\begin{align*}
  \D_1 \Phi + i \D_2 \Phi &= 0\\
  \kappa B + |\Phi|^2 &= 0.
\end{align*}
We can do exactly the same reduction as in Taubes' theorem, and we get that we need to solve
\[
  \Delta \log |\Phi| = \frac{1}{\kappa} |\Phi|^2.
\]
Alternatively, if $\rho = |\Phi|$, then
\[
  \Delta \log \rho = - \frac{2}{\kappa}\rho.
\]
This can in fact be solved explicitly.

\section{Skyrmions}
We now move on to one dimension higher, and study \emph{Skyrmions}. In recent years, there has been a lot of interests in what people call ``Skyrmions'', but what they are studying is a 2-dimensional variant of the original idea of Skyrmions. These are useful for genuine, practical things. However, we are theorists, and we do not care about such things. We are going to study the original Skyrmions as discovered by Skyrme. 

Hadronic physics exhibits (approximate) spontaneously broken \emph{chiral symmetry} $\frac{\SU(2)_L \times \SU(2)_R}{\Z/2\Z} \cong \SO(4)$, where the unbroken group is (diagonal) $\SO(3)$ isospin, and the elements of $\SO(3)$ are $(g, g) \in \SU(2) \times \SU(2)$.

This symmetry is captured in various effective field theories of pions (which are the approximate Goldstone bosons) and mesons. It is also a feature of QCD with very light $u$ and $d$ quarks.

The special feature of Skyrmion theory is that we describe nucleons as solitons in the effective field theory. Most people don't do this --- they just write the nucleons as fundamental Dirac fields and do quantum field theory in the usual way.

Skyrme's original idea was that nucleons and bigger nuclei are just classical approximations to some ``condensates'' of pion fields. To explain the conservation  of baryon numbers, the classical field equations have soliton solutions (Skyrmions) with an integer topological charge. This topological charge is then identified with what is known, physically, as the baryon number.

Let's begin by describing the detailed theory that has these solitons.
\subsection{Skyrme field and its topology}
Note that $\SU(2) \times \SU(2)$ acts naturally on $\SU(2)$ itself, by left and right multiplication. Since we want to capture this symmetry, we will construct the Skyrme fields as $\SU(2)$-valued matrices.

We let $\boldsymbol\tau$ be the Pauli matrices. Then we can write the Skyrme field as
\[
  U(\mathbf{x}, t) + \sigma(\mathbf{x}, t) \mathbf{1}_2 + i \boldsymbol\pi(x, t) \cdot \boldsymbol\tau.
\]
For this to lie in $\SU(2)$, we need the coefficients to satisfy
\[
  \sigma, \pi_i \in \R,\quad \sigma^2 + \boldsymbol\pi \cdot \boldsymbol\pi = 1.
\]
This is a \emph{non-linear} constraint. This condition is a sum of four squares. So geometrically, we are talking about the unit three sphere $S^3 \subseteq \R^4$. Indeed, $\SU(2)$ is isomorphic to the $3$-sphere. This is known as a \term{$\sigma$-model}.

Since we want a $\SU(2) \times \SU(2)$ symmetry, our Lagrangian has to be invariant under left and right multiplication of $\SU(2)$. Under the identification of $\SU(2) \cong \S^3$ geometrically, this is just the usual $\SO(4)$ action on $S^3$. If we want this $\SO(4)$ to be our isometry group, then there is a natural Riemannian metric to put on the sphere, namely the metric induced by $\R^4$.

From a physical point of view, the $\pi$ is the pion, which we have seen and observed a lot. However, as far as we can tell from experiments, there is no ``$\sigma$ meson''. This justifies our non-linear constraint $\sigma^2 + \boldsymbol\pi \cdot \boldsymbol\pi = 1$, since, up to a sign, the $\sigma$ is determined by the $\pi$ field, and is hence not a genuine new degree of freedom.

For convenience, note that
\[
  U^{-1} = \sigma \mathbf{1} - i \boldsymbol\pi \cdot \boldsymbol\tau.
\]
We adopt the $(+, -, -, -)$ metric signature, and the Skyrme Lagrangian is given by
\begin{multline*}
  L = \int \left( -\frac{F_{\pi}^2 }{16} \Tr \left( \partial_\mu UU^{-1} \partial^\mu UU^{-1}\right)\right. \\
  + \left.\frac{1}{32e^2}\Tr\left([\partial_\mu UU^{-1}, \partial_\nu UU^{-1}][\partial^\mu UU^{-1}, \partial^\nu UU^{-1}]\right) - \frac{1}{8} F_\pi^2 m_\pi^2 \Tr(\mathbf{1} - U)\right)\;\d^3 x.
\end{multline*}
Here $\partial_\mu UU^{-1}$ should be interpreted by $(\partial_\mu U)U^{-1}$, since $UU^{-1}$ is the identity and its derivative isn't that interesting. Crucially, we have $(\partial_\mu U)U^{-1} \in \su(2)$. We do need the $U^{-1}$ for this to land in the Lie algebra. % introduce R_\mu notation.

Note that by the Derrick's scaling argument, we cannot have solitons if we just have the first term, since it is quadratic in derivatives. Thus, we need the second term.

There is really only two possible term with four derivatives. The alternative is to have the square of the first term. However, Skyrme rejected that object, because that Lagrangian will have four time derivatives. From a classical point of view, this is nasty, because to specify the initial conditions, not only do we need the initial field condition, but also its first three derivatives. This doesn't happen in our theory, because the commutator vanishes when $\mu = \nu$.

These first two terms together have an exact chiral symmetry, under the transformations
\[
  U \mapsto O_1 U O_2
\]
for $O_1, O_2 \SU(2)$. Then the terms are invariant by the cyclicity of trace. In the absence of the final term this symmetry would be spontaneously broken by a choice of vacuum. There is a conventional choice $U = \mathbf{1}$. In terms of the $\sigma$ and $\pi$ fields, this choice of vacuum can be expressed as $\sigma = 1, \boldsymbol\pi = 0$. After this spontaneous symmetry breaking, we are left with an isospin $\SU(2)$ symmetry, since we require $O_1 = O_2^{-1}$. This isospin symmetry rotates the $\pi$ fields among themselves.

The role of the extra term is that the vacuum \emph{has} to be the identity matrix. The symmetry is now \emph{explicitly broken}. This might not be immediately obvious, but if you work it out, that's what it is. Note that this theory is still invariant under the isospin $\SU(2)$ symmetry. Thus, this behaves like a pion mass term. Since the isospin symmetry is not broken, all pions have the same mass. In reality, this is approximately true. The three pion fields are
\[
  \pi^{\pm} = \pi_1 \pm i \pi_2,\quad \pi^0 = \pi_3.
\]
We can attribute the difference in mass due to electromagnetic effects.

The three terms are referred to as the $\sigma$-model term, Skyrme term and pion mass term. The $\sigma$-model term and Skyrme terms are quadratic in time derivatives.

We can visualize the potential as
\begin{center}
  \begin{tikzpicture}
    \draw circle [radius=1];

    \node [circ] at (0, -1) {};
    \node [below] at (0, -1) {$\sigma = 1$};

    \node [circ] at (0, 1) {};
    \node [above] at (0, 1) {$\sigma = -1$};

    \draw [->] (2.5, -0.7) -- (2.5, 0.7) node [pos=0.5, right] {Potential energy};

    \draw [dashed] (1, 0) arc(0:180:1 and 0.3);
    \draw (1, 0) arc(0:-180:1 and 0.3);
    \node [right, align=left] at (1, 0) {$\sigma = 0$\\ $|\boldsymbol\pi| = 1$};

    \node at (-0.707, 0.707) [anchor = south east] {$S^3 \cong \SU(2)$};
  \end{tikzpicture}
\end{center}
The $\sigma = -1$ is the \emph{anti-vacuum}. We will see that in the core of the Skyrmion, it will take value $\sigma = -1$.

In the Skyrme Lagrangian, we have three free parameters. This is rather few for an effective field theory, but we are lazy and we can reduce the number by picking appropriate coefficients. We introduce an energy unit $\frac{F_\pi}{4e}$ and length unit $\frac{2}{eF_\pi}$. Setting these to $1$, there is one parameter left, which is the relative pion mass. Then in these units, we have
\[
  L = \int \left(-\frac{1}{2}\Tr(R_\mu R^\mu) + \frac{1}{16} \Tr([R_\mu, R_\nu][R^\mu, R^\nu]) - m^2 \Tr(\mathbf{1} - U)\right)\;\d^3 x,
\]
where
\[
  R_\mu = (\partial_\mu U)U^{-1}
\]
is the \term{right current}. In this notation, we have
\[
  m = \frac{2m_\pi}{e F_\pi}.
\]
In general, we will think of $m$ as being ``small''.

Let's see what happens if we in fact put $m = 0$. In this case, the lack of mass term means we no longer have the boundary condition that $U \to 1$ at infinity. Hence, we need to manually impose this condition.

Deriving the Euler--Lagrange equations is slightly messy, since we have to vary $U$ while staying inside the group $\SU(2)$. Thus, we vary $U$ multiplicatively,
\[
  U \mapsto U (1 + \varepsilon V)
\]
for some $V \in \su(2)$. We then have to figure out how $R$ varies, do some differentiation, and then the Euler--Lagrange equations turn out to be
\[
  \partial_\mu \left(R^\mu + \frac{1}{4} [R^\nu, [R_\nu, R^\mu]]\right) = 0.
\]
For static fields, the energy is given by
\[
  E = \int \left(-\frac{1}{2} \Tr(R_i R_i) - \frac{1}{16} \Tr([R_i, R_j] [R_i, R_j])\right)\;\d^3 x \equiv E_2 + E_4.
\]
where we sum $i$ and $j$ from $1$ to $3$. This is a sum of two terms --- the first involves a quadratic derivative, while the second involves a fourth derivative.

Note that the trace functional on $\su(2)$ is negative definite. So the energy is actually positive, despite what you see.

We can again run Derrick's theorem.
\begin{thm}[Derrick's theorem]
  We have $E_2 = E_4$ for any finite-energy static solution for $m = 0$ Skyrmions.
\end{thm}

\begin{proof}
  Suppose $U(\mathbf{x})$ minimizes $E = E_2 + E_4$. We rescale this solution, and define
  \[
    \tilde{U}(\mathbf{x}) = U(\lambda \mathbf{x}).
  \]
  Since $U$ is a solution, the energy is stationary with respect to $\lambda$ at $\lambda = 1$.

  We can take the derivative of this and obtain
  \[
    \partial_i \tilde{U}(\mathbf{x}) = \lambda \tilde{U}(\lambda \mathbf{x}).
  \]
  Therefore we find
  \[
    \tilde{R}_i(\mathbf{x}) = \lambda R_i(\lambda \mathbf{x}).
  \]
  So we find that we have
  \begin{align*}
    \tilde{E}_2 &= - \frac{1}{2} \int \Tr(\tilde{R}_i \tilde{R}_i) \;\d^3 x\\
    &= - \frac{1}{2} \lambda^2 \int \Tr(R_i R_i)(\lambda \mathbf{x})\;\d^3 x\\
    &= - \frac{1}{2} \frac{1}{\lambda} \int \tr(R_i R_i) (\lambda \mathbf{x})\;\d^3 (\lambda x)\\
    &= \frac{1}{\lambda} E_2.
  \end{align*}
  Similarly, we find that
  \[
    \tilde{E}_4 = \lambda E_4.
  \]
  So we find
  \[
    \tilde{E} = \frac{1}{\lambda} E_2 + \lambda E_4.
  \]
  But this function has to have a minimum at $\lambda = 1$. So the derivative with respect to $\lambda$ must vanish at $1$.

  So we obtain
  \[
    0 = \frac{\d \tilde{E}}{\d \lambda} = - \frac{1}{\lambda^2}E_2 + E_4 = 0.
  \]
  So we must have $E_4 = E_2$.
\end{proof}

We see that we must have a fourth derivative in order to stabilizer the soliton. If we have the mass term as well, then the argument is slightly more complicated, and we get some more complicated relation between the energy.

Note that the Skyrmions, which we will find later, have a definite scale size of order $1$.

\subsubsection*{Baryon number}
Recall that our field is a function $U: \R^3 \to \SU(2)$. Since we have a boundary condition $U = \mathbf{1}$ at infinity, we can imagine compactifying $\R^3$ into $S^3$, where the point at infinity is sent to $\mathbf{1}$.

On the other hand, we know that $\SU(2)$ is isomorphic to $S^3$. Geometrically, we think of these as ``different'' $S^3$. We think of $\SU(3)$ as the ``unit sphere'', and write it as $S^3_1$. However, we can think of the compactification of $\R^3$ as a spehre with ``infinite radius'', so we denote it as $S^3_\infty$. Of course, topologically, they are the same.

So this is a map
\[
  U: S^3_\infty \to \S^3_1.
\]
Thus map has a degree. There are many ways we can think about this. For example, we can think of this as the homotopy class of this map, which is an element of $\pi_3(S^3) \cong \Z$. Equivalently, we can think of this as the map induced on the homology or cohomology of $S^3$.

While $U$ evolves with time, because the degree is a discrete quantity, it has to be independent of time (alternatively, the degree of the map is homotopy invariant).

There is an explicit integral formula for the degree. We will not derive this, but it is
\[
  B = - \frac{1}{24\pi^2} \int \varepsilon_{ijk} \Tr(R_i R_j R_k) \;\d^3 x.
\]
There is a $2\pi^2$ coming from the volume of the three sphere. There is also a factor of $6$ coming from how we anti-symmetrize the term.

To obtain this, we pull back a normalized volume form from $S_1^3$ and then integrate over all space. Here we chose to use the $\SO(4)$-invariant volume form, but in general, we can pull back any volume form.

Locally, near $\sigma = 1$, this volume element is actually
\[
  \frac{1}{2\pi^2}\;\d \pi^1 \wedge\d \pi^2 \wedge\d \pi^3.
\]
\subsubsection*{Faddeev--Bogomolny bound}
There is a nice result analogous to the Bogomolny equation we saw for kinks and vortices, known as the \term{Faddeev--Bogomolny bound}. We can write the static energy as
\[
  E = \int - \frac{1}{2} \Tr\left(\left(R_i \mp \frac{1}{4} \varepsilon_{ijk}[R_j, R_k]\right)^2\right)\;\d^3 x \pm 12 \pi^2 B.
\]
This bound is true for both sign choices. However, to get the strongest result from this bound, we should choose the sign such that $\pm 12 \pi^2 B > 0$. Then we find
\[
  E \geq 12 \pi^2 |B|.  
\]
By symmetry, it suffices to consider the case $B > 0$, which is what we will do.

The Bogomolny equation should be
\[
  - \frac{1}{2} \Tr\left(R_i \mp \frac{1}{4} \varepsilon_{ijk} [R_j, R_k]\right)^2 = 0.
\]
However, it turns out this equation has \emph{no} non-vacuum solution.

Roughly, the argument goes as follows --- by careful inspection, for this to vanish, whenever $R$ is non-zero, the three vectors $R_1, R_2, R_3$ must form an orthonormal frame. So $U$ must be an isometry. But this isn't possible, because the spheres have ``different radii''.

Therefore, true Skyrmions with $B > 0$ satisfy
\[
  E > 12 \pi^2 B.
\]
We get a lower bound, but the actual energy is always greater than this lower bound. It is quite interesting to look at the energies of true solutions, and their energy is indeed bigger, if we try to do it numerically.

\subsection{Skyrmion solutions}
The simplest Skyrmion solution has baryon number $B = 1$. We will continue to set $m = 0$.
\subsubsection*{$B = 1$ hedgehog Skyrmion}
Consider the spherically symmetric function
\[
  U(\mathbf{x}) = \cos f(r) \mathbf{1} + i \sin f(r) \mathbf{x} \cdot \boldsymbol\tau.
\]
This is manifestly in $\SU(2)$, because $\cos^2 + \sin^2 = 1$. We need some boundary  conditions. We need $U \to 1$ at $\infty$. On the other hand, one can check that we need $U \to -\mathbf{1}$ at the origin to get baryon number $1$. So $f \to \pi$ as $r \to 0$, and $f \to 0$ as $r \to \infty$.




\printindex
\end{document}
